\chapter{Quadratic Programming}

Quadratic programming is similar to linear programming, but the objective function is quadratic rather than linear.
The constraints, if there are any, are still of the same form.
Thus, $G, \mathbf{h}, A$, and $\b$ are optional.
The formulation that we will use is
\begin{align*}
\text{minimize}\qquad &\frac{1}{2}\x\trp Q\x + \mathbf{r}\trp \x \\
\text{subject to}\qquad &G\x \preceq \mathbf{h}\\
 &A\x = \b,
\end{align*}
where $Q$ is a positive semidefinite symmetric matrix.
In this formulation, we require again that $A$ has full row rank and that the block matrix
$[Q \quad G \quad A]\trp $ has full column rank.

As an example, consider the quadratic function
\[
f(x_1,x_2) = 2x_1^2 +2x_1x_2 + x_2^2 +x_1 -x_2.
\]
There are no constraints, so we only need to initialize the matrix $Q$ and the vector $\mathbf{r}$.
To find these, we first rewrite our function to match the formulation given above.
If we let
\[
Q = \begin{bmatrix}
  a & b\\
  b & c
\end{bmatrix}, \qquad
\mathbf{r} = \begin{bmatrix}
  d\\
  e
\end{bmatrix},
\qquad \text{and} \qquad
\x = \begin{bmatrix}
  x_1\\
  x_2
\end{bmatrix},
\]
then
\begin{align*}
\frac{1}{2}\x\trp Q\x + \mathbf{r}\trp \x &=
\frac{1}{2}
\begin{bmatrix}
  x_1\\
  x_2
\end{bmatrix}\trp
\begin{bmatrix}
  a & b\\
  b & c
\end{bmatrix}
\begin{bmatrix}
  x_1\\
  x_2
\end{bmatrix} +
\begin{bmatrix}
  d\\
  e
\end{bmatrix}\trp
\begin{bmatrix}
  x_1\\
  x_2
\end{bmatrix} \\
&= \frac{1}{2}ax_1^2 + bx_1x_2 + \frac{1}{2}cx_2^2 + dx_1 + ex_2
\end{align*}
Thus, we see that the proper values to initialize our matrix $Q$ and vector $\mathbf{r}$ are:
\begin{align*}
a &= 4  &d = 1 \\
b &= 2  &e = -1 \\
c &= 2
\end{align*}
Now that we have the matrix $Q$ and vector $\mathbf{r}$, we are ready to use the CVXOPT function for quadratic programming \li{solvers.qp()}.
\begin{lstlisting}
>>> Q = matrix(np.array([[4., 2.], [2., 2.]]))
>>> r = matrix([1., -1.])
>>> sol=solvers.qp(Q, r)
>>> print(sol['x'])
[-1.00e+00]
[ 1.50e+00]
>>> print sol['primal objective']
-1.25
\end{lstlisting}

\begin{problem}{Quadratic Minimization}
Find the minimizer and minimum of
\begin{equation*}
g(x_1,x_2,x_3) = \frac{3}{2}x_1^2 +2x_1x_2 + x_1x_3+ 2x_2^2 +2x_2x_3+\frac{3}{2}x_3^2+3x_1 + x_3
\end{equation*}
\\(Hint: Write the function $g$ to match the formulation given above before coding.)
\begin{comment}
\begin{equation}
f(x) = \frac{1}{2}x\trp Qx - x\trp p
\end{equation}
where

\begin{center}
$Q =
\begin{bmatrix}
3 & 2 & 1\\
2 & 4 & 2\\
1 & 2 & 3\\
\end{bmatrix}
$
and $p =
\begin{bmatrix}
3\\
0\\
1\\
\end{bmatrix}
$
\end{center}
\end{comment}

\end{problem}


\begin{problem}{$l_2$ minimization}
The $l_2$ minimization problem is to
\begin{align*}
\text{minimize}\qquad &\|\x\|_2\\
\text{subject to} \qquad &A\x = \b.
\end{align*}

This problem is equivalent to a quadratic program, since $\|\x\|_2 = \x\trp \x$.
Write a function that accepts a matrix $A$ and vector $\b$ and solves the $l_2$ minimization problem.
Return the minimizer $\x$ and the primal objective value.

To test your function, use the matrix $A$ and vector $\b$ from Problem \ref{prob:l1}. 
The minimizer is approximately $\x=[0.966, 2.169, 0.809, 0.888]^T$ and the minimum primal objective value is approximately $7.079$.
\end{problem}
