\documentclass[leqno,fleqn,12pt,titlepage]{amsbook}
\usepackage{fullpage}
\usepackage{subfigure}
\usepackage{amsmath}
\usepackage{amsthm}
\usepackage{amssymb}
\usepackage{graphicx}
\usepackage{subfigure}
\usepackage{mdwlist}
\usepackage[small,bf]{caption}
\usepackage{calc}
\usepackage{float}
\usepackage{setspace} 
\usepackage{rotating}
\usepackage{verbatim}
\usepackage{textcomp}
\usepackage[pdftitle={Linear Programming Lecture Notes},
colorlinks=true,linkcolor=black,urlcolor=blue,citecolor=red]{hyperref}
\interdisplaylinepenalty=2500

%%FIX List of Figures Problem
\makeatletter
\renewcommand\l@figure{\@tocline{0}{3pt plus2pt}{0pt}{3.3em}{}}
\makeatother

%%UNCOMMENT FOR iPad notes
%\makeatletter
%\renewcommand*\@ptsize{4}
%\let\small\relax
%\let\footnotesize\relax
%\let\scriptsize\relax
%\let\tiny\relax
%\let\large\relax
%\let\Large\relax
%\let\LARGE\relax
%\let\huge\relax
%\let\Huge\relax
%\input{size1\@ptsize.clo}
%\makeatother

%\renewcommand{\thefigure}{\thechapter.\arabic{figure}}
\numberwithin{equation}{chapter}
\numberwithin{figure}{chapter}

\floatstyle{boxed}
\newfloat{cgalgorithm}{thp}{lop}
\floatname{cgalgorithm}{Algorithm}
\newenvironment{algorithm}
{\begin{cgalgorithm}[htbp]\small}
{\normalsize\end{cgalgorithm}}

%\doublespace



\title{Linear Programming: Penn State Math 484 Lecture Notes\\
		{\small Version 1.8.3}\\
		}
\author{
	Christopher Griffin\\
	\textcopyleft~2009-2014\\
	\vspace*{1em}
	{\scriptsize Licensed under a 
	\href{http://creativecommons.org/licenses/by-nc-sa/3.0/us/}
	{Creative Commons Attribution-Noncommercial-Share Alike 3.0 United States License}}
	\\
	{\small \textit{With Contributions By:}\\
	Bob Pakzad-Hurson\\ Greg Ference\\ Veselka Kafedzhieva\\ Michael Cline\\
	Akinwale Akinbiyi\\Ethan Wright\\
	Richard Benjamin\\
	Douglas Mercer\\}
	
	
}
\date{}

\newtheorem{theorem}{Theorem}[chapter]
\newtheorem{proposition}[theorem]{Proposition}
\newtheorem{lemma}[theorem]{Lemma}
\newtheorem{corollary}[theorem]{Corollary}

\theoremstyle{definition}
\newtheorem{definition}[theorem]{Definition}

\theoremstyle{remark}
\newtheorem{remark}[theorem]{Remark}
\newtheorem{example}[theorem]{Example}
\newtheorem{exercise}{Exercise}

\begin{document}
\frontmatter


\maketitle
\tableofcontents

\listoffigures


\chapter*{Preface}
This is a set of lecture notes. It is not a book. You should probably go away and come back when you have a real textbook on Linear Programming. Okay, do you have a book? Alright, let's move on then. This is a set of lecture notes for Math 484--Penn State's undergraduate Linear Programming course. Since I use these notes while I teach, there may be typographical errors that I noticed in class, but did not fix in the notes. If you see a typo, send me an e-mail and I'll add an acknowledgement. There may be many typos, that's why you should have a real textbook.

The lecture notes are (roughly) based on the first 6 chapters of Bazaraa et al.'s \textit{Linear Programming and Network Flows} book. This is a reasonably good book, written primarily by and for Industrial Engineers. The only problem I have with the book is that it does not present major results in the standard \textit{theorem-proof} style common to mathematical discourse. This set of notes corrects this problem by presenting the material in a format for presentation to a mathematics class. Many of the proofs in this set of notes are adapted from the textbook with some minor additions. (For example, each time Bazaraa et al. ask, ``Why?'' in a proof, I provide this information.) Additionally, I prefer to present maximization problems, while \textit{Linear Programming and Network Flows} prefers the minimization format. I've modified all the proofs to operate on maximization problems. When used with the book, the student can obtain a complete set of proofs for elementary Linear Programming.

In order to use these notes successfully, you should have taken courses in:
\begin{enumerate*}
\item Vector calculus (Math 230/231 at Penn State)
\item Matrix algebra (Math 220 at Penn State)
\end{enumerate*}
I review a substantial amount of the material you will need, but it's always good to have covered prerequisites before you get to a class. That being said, I hope you enjoy using these notes!
\mainmatter
\input{Section1.tex}
\input{Section2.tex}
\input{Section3.tex}
\input{Section4.tex}
\input{Section5.tex}
\input{Section6.tex}
\input{Section7.tex}
\input{Section8.tex}
\input{Section9.tex}
\backmatter
\bibliography{References.bib}
\bibliographystyle{amsalpha}
\end{document}
\end
