\begin{enumialphparenastyle}

\begin{ex} The wind blows from the South at $20$ kilometers per hour and an
airplane which flies at $600$ kilometers per hour in still air is heading
East. Find the velocity of the airplane and its location after
two hours. \vspace{1mm}
%\begin{sol}
%\end{sol}
\end{ex}

\begin{ex} The wind blows from the West at $30$ kilometers per hour and an
airplane which flies at $400$ kilometers per hour in still air is heading
North East. Find the velocity of the airplane and its position
after two hours. \vspace{1mm}
%\begin{sol}
%\end{sol}
\end{ex}

\begin{ex} The wind blows from the North at $10$ kilometers per hour. An
airplane which flies at $300$ kilometers per hour in still air is supposed
to go to the point whose coordinates are at $\left( 100, 100 \right). $ In what direction should the airplane fly? \vspace{1mm}
%\begin{sol}
%\end{sol}
\end{ex}

\begin{ex} Three forces act on an object. Two are $\leftB
\begin{array}{r}
3 \\
-1 \\
-1
\end{array}
\rightB $ and $\leftB
\begin{array}{r}
1 \\
-3 \\
4
\end{array}
\rightB $ Newtons. Find the third force if the object is not to move.
\vspace{1mm}
%\begin{sol}
%\end{sol}
\end{ex}

\begin{ex} Three forces act on an object. Two are $\leftB
\begin{array}{r}
6 \\
-3 \\
3
\end{array}
\rightB $ and $\leftB
\begin{array}{r}
2 \\
1 \\
3
\end{array}
\rightB $ Newtons. Find the third force if the total force on the object is
to be $\leftB
\begin{array}{r}
7 \\
1 \\
3
\end{array}
\rightB .$ \vspace{1mm}
%\begin{sol}
%\end{sol}
\end{ex}

\begin{ex} A river flows West at the rate of $b$ miles per hour. A boat can move
at the rate of $8$ miles per hour. Find the smallest value of $b$ such that
it is not possible for the boat to proceed directly across the river.
\vspace{1mm}
%\begin{sol}
%\end{sol}
\end{ex}

\begin{ex} The wind blows from West to East at a speed of $50$ miles per hour and
an airplane which travels at $400$ miles per hour in still air is heading
North West. What is the velocity of the airplane relative to the ground?
What is the component of this velocity in the direction North? \vspace{1mm}
\begin{sol}
The velocity is the sum of two vectors. $50\vect{i}+\frac{
300}{\sqrt{2}} \left( \vect{i}+\vect{j}\right) =\left( 50+\frac{300}{\sqrt{2}}
\right) \vect{i}+ \frac{300}{\sqrt{2}}\vect{j}.$ The component in the
direction of North is then $\frac{300}{\sqrt{2}}= 150\sqrt{2}$
and the velocity relative to the ground is
\[
\left( 50+\frac{300}{\sqrt{2}}\right) \vect{i}+\frac{300}{\sqrt{2}}\vect{j}
\]
\end{sol}
\end{ex}

\begin{ex} The wind blows from West to East at a speed of $60$ miles per hour and
an airplane can travel travels at $100$ miles per hour in still air. How
many degrees West of North should the airplane head in order to travel
exactly North? \vspace{1mm}
%\begin{sol}
%\end{sol}
\end{ex}


\begin{ex} The wind blows from West to East at a speed of $50$ miles per hour and
an airplane which travels at $400$ miles per hour in still air heading
somewhat West of North so that, with the wind, it is flying due North. It
uses $30.0$ gallons of gas every hour. If it has to travel $600.0$ miles due
North, how much gas will it use in flying to its destination? \vspace{1mm}
%\begin{sol}
%\end{sol}
\end{ex}

\begin{ex} An airplane is flying due north at $150.0$ miles per hour but it is
not actually going due North because there is a wind which is pushing the
airplane due east at $40.0$ miles per hour. After one hour, the plane starts
flying $30^{\circ }$ East of North. Assuming the plane starts at $\left(
0,0\right) ,$ where is it after $2$ hours? Let North be the direction of the
positive $y$ axis and let East be the direction of the positive $x$ axis.
\vspace{1mm}
\begin{sol}
 Velocity of plane for the first hour: $\leftB \begin{array}{cc} 
0 & 150
\end{array}
\rightB  + \leftB \begin{array}{cc}
40 & 0
\end{array}
\rightB =\leftB
\begin{array}{cc}
40 & 150
\end{array}
\rightB .$ After one hour it is at $\left( 40,150\right) .$ Next the
velocity of the plane is $150\leftB \begin{array}{cc}
 \frac{1}{2} & \frac{\sqrt{3}}{2}
\end{array}
\rightB
+\leftB
\begin{array}{cc}
 40 & 0
\end{array}
\rightB $ in miles per hour. After two hours it is then at 
$\left( 40,150\right) + 150\leftB \begin{array}{cc}
 \frac{1}{2} & \frac{\sqrt{3}}{2}
\end{array}
\rightB
+\leftB
\begin{array}{cc}
 40 & 0
\end{array}
\rightB  =  \leftB
\begin{array}{cc}
155 & 75\sqrt{3}+150
\end{array}
\rightB = \leftB
\begin{array}{cc}
155.0 & 279.\,\allowbreak 9
\end{array}
\rightB $
\end{sol}
\end{ex}

\begin{ex} 
City A is located at the origin $\left( 0,0 \right)$ while city B is located at $\left(300,500 \right) $ where distances are in miles. An airplane flies at $250$
miles per hour in still air. This airplane wants to fly from city A to city
B but the wind is blowing in the direction of the positive $y$ axis at a
speed of $50$ miles per hour. Find a unit vector such that if the plane heads
in this direction, it will end up at city B having flown the shortest
possible distance. How long will it take to get there? \vspace{1mm}
\begin{sol}
Wind: $\leftB \begin{array}{cc}
0 & 50
\end{array}
\rightB .$ Direction it needs to travel: $\left( 3,5 \right) \frac{1}{\sqrt{34}}.$ Then you need $250 \leftB \begin{array}{cc}
 a & b
\end{array}
\rightB + \leftB \begin{array}{cc}
0 & 50
\end{array}
\rightB $ to
have this direction where $\leftB 
\begin{array}{cc}
 a & b
\end{array}
\rightB $ is an appropriate unit
vector. Thus you need
\begin{eqnarray*}
a^{2}+b^{2} &=&1 \\ 
& & \\
\frac{250b+50}{250a} &=&\frac{5}{3}
\end{eqnarray*}
Thus $a=\frac{3}{5},b=\frac{4}{5}.$ The velocity of the plane relative
to the ground is $\leftB
\begin{array}{cc}
150 & 250
\end{array}
\rightB .$ The speed of the plane relative to the ground is given by 
\[
\sqrt{\left( 150\right) ^{2}+\left( 250\right) ^{2}}=
291.55 \text{ miles per hour}
\]
It has to go a distance of $\sqrt{\left( 300\right) ^{2}+\left( 500\right)
^{2}}=\allowbreak 583.\,\allowbreak 10$ miles. Therefore, it takes
\[
\frac{\allowbreak 583.\,\allowbreak 1}{\allowbreak 291.\,\allowbreak 55}=2
\text{ hours}
\]
\end{sol}
\end{ex}

\begin{ex} A certain river is one half mile wide with a current flowing at $2$
miles per hour from East to West. A man swims directly toward the
opposite shore from the South bank of the river at a speed of $3$ miles
per hour. How far down the river does he find himself when he has swam
across? How far does he end up traveling? \vspace{1mm}
\begin{sol}
Water:$\leftB \begin{array}{rr}
-2 & 0
\end{array}
\rightB $ 
Swimmer:$\leftB \begin{array}{rr}
0 & 3
\end{array}
\rightB $ 
Speed relative to earth: $\leftB
\begin{array}{rr}
 -2 & 3
\end{array}
\rightB .$ 
It takes him $1/6$ of an hour to get across. Therefore, he ends up travelling $\frac{1}{6}\sqrt{4+9}= \frac{1}{6}\sqrt{13}$ miles. He ends up $1/3$ mile
down stream.
\end{sol}
\end{ex}

\begin{ex} A certain river is one half mile wide with a current flowing at 2
miles per hour from East to West. A man can swim at $3$ miles per hour in
still water. In what direction should he swim in order to travel directly
across the river? What would the answer to this problem be if the river
flowed at 3 miles per hour and the man could swim only at the rate of 2
miles per hour? \vspace{1mm}
\begin{sol}
Man:\ $3\leftB \begin{array}{rr}
 a & b
\end{array}
\rightB $ 
Water:\ $\leftB \begin{array}{rr}
-2 & 0
\end{array}
\rightB $
Then you need $3a=2$ and so $a=2/3$ and hence $b=\sqrt{5}/3$. The vector is
then $\leftB \begin{array}{cc}
 \frac{2}{3} & \frac{\sqrt{5}}{3}
\end{array}
\rightB .$ 

In the second case, he
could not do it. You would need to have a unit vector $\leftB \begin{array}{rr} a & b
\end{array}
\rightB $
such that $2a=3$ which is not possible. 
\end{sol}
\end{ex}


\begin{ex} Three forces are applied to a point which does not move. Two of the
forces are $2 \vect{i}+2 \vect{j} -6 \vect{k}$ Newtons and $8 \vect{i}+ 8 \vect{j}+ 3 \vect{k}$ Newtons. Find
the third force. \vspace{1mm}
%\begin{sol}
%\end{sol}
\end{ex}

\begin{ex} The total force acting on an object is to be $4 \vect{i}+
 2 \vect{j} -3 \vect{k}$ Newtons. A force
of $-3 \vect{i} -1 \vect{j}+ 8
\vect{k}$ Newtons is being applied. What other force should be applied to
achieve the desired total force? \vspace{1mm}
%\begin{sol}
%\end{sol}
\end{ex}

\begin{ex} A bird flies from its nest $8$ km in the direction $\frac{5}{6}\pi $
north of east where it stops to rest on a tree. It then flies $1$ km in the
direction due southeast and lands atop a telephone pole. Place an $xy$
coordinate system so that the origin is the bird's nest, and the positive 
$x$ axis points east and the positive $y$ axis points north. Find the
displacement vector from the nest to the telephone pole. \vspace{1mm}
%\begin{sol}
%\end{sol}
\end{ex}


\begin{ex} If $\vect{F}$ is a force and $\vect{D}$ is a vector, show 
$\func{proj}_{\vect{D}}\left( \vect{F}\right) =\left( \vectlength \vect{F}
\vectlength \cos \theta \right) \vect{u}$ where $\vect{u}$ is the unit
vector in the direction of $\vect{D}$, where $\vect{u}=\vect{D}/\vectlength \vect{D}
\vectlength $ and $\theta $ is the included angle between the two vectors, 
$\vect{F}$ and $\vect{D}$. $\vectlength \vect{F}\vectlength \cos \theta $
is sometimes called the component
\index{component of a force} of the force, $\vect{F}$ in the direction, $\vect{D}$.
\begin{sol}
$\func{proj}_{\vect{D}}\left( \vect{F}\right) = \frac{\vect{F}\dotprod
\vect{D}}{\vectlength \vect{D}\vectlength }\frac{\vect{D}}{\vectlength \vect{D}
\vectlength }=\left( \vectlength \vect{F}\vectlength \cos \theta \right)
\frac{\vect{D}}{\vectlength \vect{D}\vectlength }=\left( \vectlength
\vect{F}\vectlength \cos \theta \right) \vect{u}$
\end{sol}
\end{ex}

\end{enumialphparenastyle}