\begin{enumialphparenastyle}

\begin{ex} Consider the following scalar equation of a plane. 
\begin{equation*}
2x-3y+z=0
\end{equation*}
Find the orthogonal complement of the vector $\vect{v}=\leftB \begin{array}{r}
3\\
4 \\
1
\end{array}
\rightB $.
Also find the point on the plane which is closest to $\left( 3,4,1\right) .$
%\begin{sol}
%\end{sol}
\end{ex}


\begin{ex} Consider the following scalar equation of a plane. 
\begin{equation*}
x+3y+z=0
\end{equation*}
Find the orthogonal complement of the vector $\vect{v}=\leftB
\begin{array}{r}
 1 \\
2 \\
1
\end{array}
\rightB $.
Also find the point on the plane which is closest to $\left( 3,4,1\right) .$
%\begin{sol}
%\end{sol}
\end{ex}

\begin{ex} Let $\vect{v}$ be a vector and let $\vect{n}$ be a normal vector for a
plane through the origin. Find the equation of the line through the point
determined by $\vect{v}$ which has direction vector $\vect{n}.$ Show that it
intersects the plane at the point determined by $\vect{v}-\func{proj}_{\vect{n}
}\vect{v}$. \ \textbf{Hint: }The line:\ $\vect{v}+t\vect{n}.$ It is in the
plane if $\vect{n}\dotprod \left( \vect{v}+t\vect{n}\right) =0.$ Determine $t$.
Then substitute in to the equation of the line.
%\begin{sol}
%\end{sol}
\end{ex}

\begin{ex} As shown in the above problem, one can find the closest point to $\vect{v}$ in a plane through the origin by finding the intersection of the line
through $\vect{v}$ having direction vector equal to the normal vector to the
plane with the plane. If the plane does not pass through the origin, this
will still work to find the point on the plane closest to the point
determined by $\vect{v}$. Here is a relation which defines a plane 
\begin{equation*}
2x+y+z=11
\end{equation*}
and here is a point:\ $\left( 1,1,2\right) $. Find the point on the plane
which is closest to this point. Then determine the distance from the point
to the plane by taking the distance between these two points. \textbf{Hint: }
Line: $\left( x,y,z\right) =\left( 1,1,2\right) +t\left( 2,1,1\right) .$ Now
require that it intersect the plane. $\ $
%\begin{sol}
%\end{sol}
\end{ex}

\begin{ex} In general, you have a point $\left( x_{0},y_{0},z_{0}\right) $ and a
scalar equation for a plane $ax+by+cz=d$ where $a^{2}+b^{2}+c^{2}>0$.
Determine a formula for the closest point on the plane to the given point.
Then use this point to get a formula for the distance from the given point
to the plane. \textbf{Hint: }Find the line perpendicular to the plane which
goes through the given point: $\left( x,y,z\right) =\left(
x_{0},y_{0},z_{0}\right) +t\left( a,b,c\right) .$ Now require that this
point satisfy the equation for the plane to determine $t$.
%\begin{sol}
%\end{sol}
\end{ex}

\end{enumialphparenastyle}