\begin{enumialphparenastyle}

\begin{ex} If $A$ is the matrix of a linear transformation which rotates all
vectors in $\mathbb{R}^{2}$ through $60^{\circ },$ explain why $A$ cannot
have any real eigenvalues. Is there an angle such that rotation through this
angle would have a real eigenvalue? What eigenvalues would be obtainable in
this way? \vspace{1mm} 
%\begin{sol}
%\end{sol}
\end{ex}


\begin{ex} Let $A$ be the $2\times 2$ matrix of the linear transformation which
rotates all vectors in $\mathbb{R}^{2}$ through an angle of $\theta $. For
which values of $\theta $ does $A$ have a real eigenvalue? 
\begin{sol}
When you think of this geometrically, it is clear that the only two values
of $\theta $ are 0 and $\pi $ or these added to integer multiples of $2\pi $
\end{sol}
\end{ex}


\begin{ex} Let $T\,$\ be the linear transformation which reflects vectors about
the $x$ axis. Find a matrix for $T$ and then find its eigenvalues and
eigenvectors. 
\begin{sol}
The matrix of $T$ is $\leftB
\begin{array}{rr}
1 & 0 \\
0 & -1
\end{array}
\rightB$. The eigenvectors and eigenvalues are:
\[
\left\{ \leftB
\begin{array}{c}
0 \\
1
\end{array}
\rightB \right\} \leftrightarrow -1,\left\{ \leftB
\begin{array}{c}
1 \\
0
\end{array}
\rightB \right\} \leftrightarrow 1
\]
\end{sol}
\end{ex}

\begin{ex} Let $T\,$be the linear transformation which rotates all vectors in 
$\mathbb{R}^{2}$ counterclockwise through an angle of $\pi /2.$ Find a matrix
of $T$ and then find eigenvalues and eigenvectors.
\begin{sol}
The matrix of $T$ is $\leftB
\begin{array}{rr}
0 & -1 \\
1 & 0
\end{array}
\rightB$. The eigenvectors and eigenvalues are:
\[
\left\{ \leftB
\begin{array}{r}
-i \\
1
\end{array}
\rightB \right\} \leftrightarrow -i,\left\{ \leftB
\begin{array}{c}
i \\
1
\end{array}
\rightB \right\} \leftrightarrow i
\]
\end{sol}
\end{ex}

\begin{ex} Let $T$ be the linear transformation which reflects all vectors in $
\mathbb{R}^{3}$ through the $xy$ plane. Find a matrix for $T$ and then
obtain its eigenvalues and eigenvectors. 
\begin{sol}
The matrix of $T$ is $\leftB
\begin{array}{rrr}
1 & 0 & 0 \\
0 & 1 & 0 \\
0 & 0 & -1
\end{array}
\rightB$
The eigenvectors and eigenvalues are:
\[
\left\{ \leftB
\begin{array}{c}
0 \\
0 \\
1
\end{array}
\rightB \right\} \leftrightarrow -1,\left\{ \leftB
\begin{array}{c}
1 \\
0 \\
0
\end{array}
\rightB ,\leftB
\begin{array}{c}
0 \\
1 \\
0
\end{array}
\rightB \right\} \leftrightarrow 1
\]
\end{sol}
\end{ex}

\end{enumialphparenastyle}