\begin{enumialphparenastyle}

\begin{ex}
Find the point $\left( x_{1},y_{1}\right) $ which
lies on both lines, $x+3y=1$ and $4x-y=3.$
\begin{sol}
$\begin{array}{c}
x+3y=1 \\
4x-y=3
\end{array}
$, Solution is: $\left[ x=\frac{10}{13},y=\frac{1}{13}\right] $
\end{sol}
\end{ex}

\begin{ex}
Find the point of intersection of the two lines $
3x+y=3$ and $x+2y=1.$ 
\begin{sol}
$\begin{array}{c}
x+3y=1 \\
4x-y=3
\end{array}
$, Solution is: $\left[ x=\frac{10}{13},y=\frac{1}{13}\right] $
\end{sol}
\end{ex}


\begin{ex}
Do the three lines, $x+2y=1,2x-y=1,$ and $4x+3y=3$ have a common point
of intersection? If so, find the point and if not, tell why they don't have
such a common point of intersection. 
\begin{sol}
$\begin{array}{c}
x+2y=1 \\
2x-y=1 \\
4x+3y=3
\end{array}
$, Solution is: $\left[ x=\frac{3}{5},y=\frac{1}{5}\right] $
\end{sol}
\end{ex}

\begin{ex}
Do the three planes, $x+y-3z=2,$ $2x+y+z=1,$ and $3x+2y-2z=0$ have a
common point of intersection? If so, find one and if not, tell why there is
no such point. 
\begin{sol}
No solution exists. You can see this by writing the augmented
matrix and doing row operations. $\leftB
\begin{array}{rrrr}
1 & 1 & -3 & 2 \\
2 & 1 & 1 & 1 \\
3 & 2 & -2 & 0
\end{array}
\rightB $, row echelon form: $\leftB
\begin{array}{rrrr}
1 & 0 & 4 & 0 \\
0 & 1 & -7 & 0 \\
0 & 0 & 0 & 1
\end{array}
\rightB .$ Thus one of the equations says $0=1$ in an equivalent system of
equations.
\end{sol}
\end{ex}


\begin{ex}
Four times the weight of Gaston is 150 pounds more than the weight of
Ichabod. Four times the weight of Ichabod is 660 pounds less than seventeen
times the weight of Gaston. Four times the weight of Gaston plus the weight
of Siegfried equals 290 pounds. Brunhilde would balance all three of the
others. Find the weights of the four people.
\begin{sol}
$\begin{array}{c}
4g-I=150 \\
4I-17g=-660 \\
4g+s=290 \\
g+I+s-b=0
\end{array}
$, Solution is : $\left\{ g=60,I=90,b=200,s=50\right\} $
\end{sol}
\end{ex}

\end{enumialphparenastyle}