\Opensolutionfile{solutions}[ex]
\section*{Exercises}

\begin{enumialphparenastyle}

\begin{ex} Suppose you have $\mathbb{R}^{2}$ and the $+$ operation is as
follows:\ 
\begin{equation*}
\left( a,b\right) +\left( c,d\right) =\left( a+d,b+c\right) .
\end{equation*}
Scalar multiplication is defined in the usual way. Is this a vector space?
Explain why or why not. 
%\begin{sol}
%\end{sol}
\end{ex}

\begin{ex} Suppose you have $\mathbb{R}^{2}$ and the $+$ operation is defined as
follows. 
\begin{equation*}
\left( a,b\right) +\left( c,d\right) =\left( 0,b+d\right)
\end{equation*}
Scalar multiplication is defined in the usual way. Is this a vector space?
Explain why or why not. 
%\begin{sol}
%\end{sol}
\end{ex}

\begin{ex} Suppose you have $\mathbb{R}^{2}$ and scalar multiplication is defined
as $c\left( a,b\right) =\left( a,cb\right) $ while vector addition is
defined as usual. Is this a vector space? Explain why or why not.
%\begin{sol}
%\end{sol}
\end{ex}

\begin{ex} Suppose you have $\mathbb{R}^{2}$ and the $+$ operation is defined as
follows. 
\begin{equation*}
\left( a,b\right) +\left( c,d\right) =\left( a-c,b-d\right)
\end{equation*}
Scalar multiplication is same as usual. Is this a vector space? Explain why
or why not. 
%\begin{sol}
%\end{sol}
\end{ex}

\begin{ex} \label{functions}Consider all the functions defined on a non empty set
which have values in $\mathbb{R}$. Is this a vector space? Explain.
The operations are defined as follows. Here $f,g$ signify functions and $a$
is a scalar. 
\begin{eqnarray*}
\left( f+g\right) \left( x\right) &=&f\left( x\right) +g\left( x\right) \\
\left( af\right) \left( x\right) &=&a\left( f\left( x\right) \right)
\end{eqnarray*}
%\begin{sol}
%\end{sol}
\end{ex}


\begin{ex} Denote by $\mathbb{R}^{\mathbb{N}}$ the set of real valued sequences.
For $\vect{a}\equiv \left\{ a_{n}\right\} _{n=1}^{\infty },\vect{b}\equiv
\left\{ b_{n}\right\} _{n=1}^{\infty }$ two of these, define their sum to be
given by 
\begin{equation*}
\vect{a}+\vect{b} =  \left\{ a_{n}+b_{n}\right\} _{n=1}^{\infty }
\end{equation*}
and define scalar multiplication by 
\begin{equation*}
c\vect{a}=\left\{ ca_{n}\right\} _{n=1}^{\infty }\text{ where }\vect{a}
=\left\{ a_{n}\right\} _{n=1}^{\infty }
\end{equation*}
Is this a special case of Problem \ref{functions}? Is this a vector space?
%\begin{sol}
%\end{sol}
\end{ex}

\begin{ex} Let $\mathbb{C}^{2}$ be the set of ordered pairs of complex numbers.
Define addition and scalar multiplication in the usual way.
\begin{equation*}
\left( z,w\right) +\left( \hat{z},\hat{w}\right) = \left( z+\hat{z},w+
\hat{w}\right) ,\ u\left( z,w\right) \equiv \left( uz,uw\right)
\end{equation*}
Here the scalars are from $\mathbb{C}$. Show this is a vector space.
%\begin{sol}
%\end{sol}
\end{ex}

\begin{ex} Let $V$ be the set of functions defined on a nonempty set which have
values in a vector space $W.$ Is this a vector space? Explain.
%\begin{sol}
%\end{sol}
\end{ex}

\begin{ex} Consider the space of $m\times n$ matrices with operation of addition
and scalar multiplication defined the usual way. That is, if $A,B$ are two $
m\times n$ matrices and $c$ a scalar, 
\begin{equation*}
\left( A+B\right) _{ij}=A_{ij}+B_{ij},\ \left( cA\right) _{ij}\equiv c\left(
A_{ij}\right)
\end{equation*}
%\begin{sol}
%\end{sol}
\end{ex}

\begin{ex} Consider the set of $n\times n$ symmetric matrices. That is, $A=A^{T}.$
In other words, $A_{ij}=A_{ji}$. Show that this set of symmetric matrices is
a vector space and a subspace of the vector space of $n\times n$ matrices.
%\begin{sol}
%\end{sol}
\end{ex}

\begin{ex} Consider the set of all vectors in $\mathbb{R}^{2},\left( x,y\right) $
such that $x+y\geq 0.$ Let the vector space operations be the usual ones. Is
this a vector space? Is it a subspace of $\mathbb{R}^{2}$?
%\begin{sol}
%\end{sol}
\end{ex}

\begin{ex} Consider the vectors in $\mathbb{R}^{2},\left( x,y\right) $ such that $xy=0$. Is this a subspace of $\mathbb{R}^{2}?$ Is it a vector space? The
addition and scalar multiplication are the usual operations.
%\begin{sol}
%\end{sol}
\end{ex}

\begin{ex} Define the operation of vector addition on $\mathbb{R}^{2}$ by $\left(
x,y\right) +\left( u,v\right) =\left( x+u,y+v+1\right) .$ Let scalar
multiplication be the usual operation. Is this a vector space with these
operations? Explain.
%\begin{sol}
%\end{sol}
\end{ex}

\begin{ex} Let the vectors be real numbers. Define vector space operations in the
usual way. That is $x+y$ means to add the two numbers and $xy$ means to
multiply them. Is $\mathbb{R}$ with these operations a vector space? Explain.
%\begin{sol}
%\end{sol}
\end{ex}

\begin{ex} Let the scalars be the rational numbers and let the vectors be
real numbers which are the form $a+b\sqrt{2}$ for $a,b$ rational numbers.
Show that with the usual operations, this is a vector space.
%\begin{sol}
%\end{sol}
\end{ex}

\begin{ex} Let $\mathbb{P}_{2}$ be the set of all polynomials of degree 2 or
less. That is, these are of the form $a+bx+cx^{2}$. Addition is defined as 
\begin{equation*}
\left( a+bx+cx^{2}\right) +\left( \hat{a}+\hat{b}x+\hat{c}x^{2}\right)
=\left( a+\hat{a}\right) +\left( b+\hat{b}\right) x+\left( c+\hat{c}\right)
x^{2}
\end{equation*}
and scalar multiplication is defined as 
\begin{equation*}
d\left( a+bx+cx^{2}\right) =da+dbx+cdx^{2}
\end{equation*}
Show that, with this definition of the vector space operations that $\mathbb{
P}_{2}$ is a vector space. Now let $V$ denote those polynomials $a+bx+cx^{2}$
such that $a+b+c=0$. Is $V$ a subspace of $\mathbb{P}_{2}?$ Explain.
%\begin{sol}
%\end{sol}
\end{ex}

\begin{ex} Let $M,N$ be subspaces of a vector space $V$ and consider $M+N$
defined as the set of all $m+n$ where $m\in M$ and $n\in N$. Show that $M+N$
is a subspace of $V$.
%\begin{sol}
%\end{sol}
\end{ex}

\begin{ex} Let $M,N$ be subspaces of a vector space $V$. Then $M\cap N$ consists
of all vectors which are in both $M$ and $N$. Show that $M\cap N$ is a
subspace of $V$.
%\begin{sol}
%\end{sol}
\end{ex}

\begin{ex} Let $M,N$ be subspaces of a vector space $\mathbb{R}^{2}.$ Then $N\cup
M$ consists of all vectors which are in either $M$ or $N$. Show that $N\cup
M $ is not necessarily a subspace of $\mathbb{R}^{2}$ by giving an example
where $N\cup M$ fails to be a subspace.
%\begin{sol}
%\end{sol}
\end{ex}


\begin{ex} \label{4julyprob1}Let $X$ consist of the real valued functions which
are defined on an interval $\left[ a,b\right] .$ For $f,g\in X,\;f+g$ is the
name of the function which satisfies $\left( f+g\right) \left( x\right)
=f\left( x\right) +g\left( x\right)$. For $s$ a real number, $
\left( s f\right) \left( x\right) = s \left( f\left( x\right)
\right) $. Show this is a vector space. 
\begin{sol}
The axioms of a vector space all hold because they
hold for a vector space. The only thing left to verify is the
assertions about the things which are supposed to exist. $0$ would
be the zero function which sends everything to $0$. This is an additive
identity. Now if $f$ is a function, $-f\left( x\right) \equiv \left(
-f\left( x\right) \right) $. Then
\[
\left( f+\left( -f\right) \right) \left( x\right) \equiv f\left( x\right)
+\left( -f\right) \left( x\right) \equiv f\left( x\right) +\left( -f\left(
x\right) \right) =0
\]
Hence $f+-f=\mathbf{0}$. For each $x\in \left[ a,b\right] ,$ let $%
f_{x}\left( x\right) =1$ and $f_{x}\left( y\right) =0$ if $y\neq x.$ Then
these vectors are obviously linearly independent.
\end{sol}
\end{ex}

\begin{ex} Consider functions defined on $\left\{ 1,2,\cdots ,n\right\} $ having
values in $\mathbb{R}$. Explain how, if $V$ is the set of all such
functions, $V$ can be considered as $\mathbb{R}^{n}$.
\begin{sol}
Let $f\left( i\right) $ be the $i^{th}$ component of a vector $
\vect{x}\in \mathbb{R}^{n}$. Thus a typical element in $\mathbb{R}^{n}$ is $
\left( f\left( 1\right) ,\cdots ,f\left( n\right) \right) $.
\end{sol}
\end{ex}

\begin{ex} Let the vectors be polynomials of degree no more than 3. Show that
with the usual definitions of scalar multiplication and addition wherein,
for $p\left( x\right) $ a polynomial, $\left( a p\right) \left(
x\right) = a p\left( x\right) $ and for $p,q$ polynomials $\left(
p+q\right) \left( x\right) =  p\left( x\right) +q\left( x\right) ,$ this
is a vector space.
\begin{sol}
This is just a subspace of the vector space of functions
because it is closed with respect to vector addition and scalar
multiplication. Hence this is a vector space.
\end{sol}
\end{ex}

\end{enumialphparenastyle}