% Appendix: Optimization Software Resources
% This appendix provides a comprehensive list of optimization software

\chapter{Optimization Software Resources}
\label{app:software}

This appendix provides a comprehensive list of optimization software packages organized by problem type. Many commercial solvers offer free academic licenses, making them accessible for educational and research purposes.

\section{Software by Problem Type}

\begin{table}[H]
\centering
\begin{tabular}{lp{10cm}}
\hline
\textbf{Category} & \textbf{Software (License)} \\ \hline
Linear programming (free) & CDD, CLP, GLPK, LPSOLVE, OSQP, SCIP \\[0.5em]
Mixed Integer Linear programming (free) & CBC, GLPK, LPSOLVE, SCIP \\[0.5em]
Linear programming (commercial) & COPT (free for academia), CPLEX (free for academia), GUROBI (free for academia), KNITRO, LINPROG, MOSEK (free for academia), XPRESS (free for academia) \\[0.5em]
Mixed Integer Linear programming (commercial) & COPT (free for academia), CPLEX (free for academia), GUROBI (free for academia), INTLINPROG, KNITRO, MOSEK (free for academia), XPRESS (free for academia) \\[0.5em]
Quadratic programming (free) & OSQP, CLP, DAQP (mixed-binary), OOQP, QPC, QPOASES, QUADPROGBB (nonconvex QP) \\[0.5em]
Quadratic programming (commercial) & COPT (free for academia), CPLEX (free for academia), GUROBI (free for academia), KNITRO, MOSEK (free for academia), QUADPROG, XPRESS (free for academia) \\[0.5em]
Mixed Integer Quadratic programming (commercial) & COPT (free for academia), CPLEX (free for academia), GUROBI (free for academia), KNITRO, MOSEK (free for academia), XPRESS (free for academia) \\[0.5em]
Second-order cone programming (free) & ECOS, SCS, SDPT3, SEDUMI \\[0.5em]
Second-order cone programming (commercial) & COPT (free for academia), CPLEX (free for academia), CONEPROG, GUROBI (free for academia), MOSEK (free for academia), XPRESS (free for academia) \\[0.5em]
Mixed Integer Second-order cone programming (commercial) & COPT (free for academia), CPLEX (free for academia), GUROBI (free for academia), MOSEK (free for academia), XPRESS (free for academia) \\[0.5em]
Semidefinite programming (free) & CSDP, DSDP, LOGDETPPA, PENLAB, SCS, SDPA, SDPLR, SDPT3, SDPNAL, SEDUMI \\[0.5em]
Semidefinite programming (commercial) & COPT (free for academia), LMIlab (not recommended), MOSEK (free for academia), PENBMI, PENSDP (free for academia) \\[0.5em]
General nonlinear programming and other solvers & BARON, FILTERSD, FMINCON, GPPSOY, IPOPT, KNITRO, LMIrank, MPT, NOMAD, PENLAB, SNOPT, SPARSEPOP \\ \hline
\end{tabular}
\caption{Optimization Software and their Licenses}
\label{tab:optimization_software_appendix}
\end{table}

\section{Recommended Solvers for This Book}

For the examples in this book, we primarily use:

\begin{description}
\item[Gurobi:] A high-performance commercial solver with excellent support for LP, MIP, and QP problems. Free for academic use with registration at \url{https://www.gurobi.com/academia/academic-program-and-licenses/}.

\item[GLPK (GNU Linear Programming Kit):] A free, open-source solver for LP and MIP problems. Available at \url{https://www.gnu.org/software/glpk/}.

\item[CBC (COIN-OR Branch and Cut):] A free, open-source MIP solver. Part of the COIN-OR project at \url{https://github.com/coin-or/Cbc}.

\item[SCIP:] One of the fastest non-commercial solvers for MIP. Free for academic use at \url{https://www.scipopt.org/}.
\end{description}

\section{Modeling Languages and Interfaces}

In addition to solvers, you'll need a way to formulate your optimization problems. We use:

\begin{description}
\item[Python with PuLP:] A free, open-source library for formulating LP and MIP problems in Python. Install with \texttt{pip install pulp}.

\item[Python with Gurobipy:] Gurobi's native Python interface, providing direct access to all Gurobi features. Included with Gurobi installation.

\item[Excel Solver:] Built into Microsoft Excel, suitable for small problems and educational purposes.

\item[AMPL:] A widely used algebraic modeling language. The AMPL Book is available for free download at \url{https://ampl.com/resources/the-ampl-book/chapter-downloads/}.

\item[AIMMS:] A commercial modeling system with free academic licenses. Optimization modeling resources available at \url{https://www.aimms.com/english/developers/resources/manuals/optimization-modeling/}.
\end{description}

\section{Gurobi Tutorials and Resources}

Gurobi provides extensive free educational materials:

\begin{itemize}
\item Example files and code: \url{https://github.com/Gurobi}
\item \href{https://www.gurobi.com/resource/mathematical-programming-tutorial-linear-programming/}{Linear Programming Tutorial}
\item \href{https://www.gurobi.com/resource/tutorial-mixed-integer-linear-programming/}{Mixed Integer Linear Programming Tutorial}
\item \href{https://www.gurobi.com/resource/python-i-webinar/}{Python I: Modeling with Gurobi in Python}
\item \href{https://www.gurobi.com/resource/python-ii-webinar/}{Python II: Advanced Algebraic Modeling}
\item \href{https://www.gurobi.com/resource/python-iii-webinar/}{Python III: Optimization and Heuristics}
\item \href{https://www.gurobi.com/resources/?category-filter=tutorials}{All Gurobi Tutorials}
\end{itemize}
