%By Kevin Cheung
%The book is licensed under the
%\href{http://creativecommons.org/licenses/by-sa/4.0/}{Creative Commons
%Attribution-ShareAlike 4.0 International License}.
%
%This file has been modified by Robert Hildebrand 2020.  
%CC BY SA 4.0 licence still applies.

\chapter{Using an LP solver}\label{using-an-lp-solver}

Linear programming problems are rarely solved by hand. Problems from
industrial applications often have thousands (and sometimes millions) of
variables and constraints. Fortunately, there exist a number of
commercial as well as open-source solvers that can handle such
large-scale problem. In this chapter, we will take a look at one of
them. Before we do so, we introduce the CPLEX LP format which allows us
to specify linear programming problems in a way close to how we write
them.

\section{CPLEX LP file format}\label{cplex-lp-file-format}

A description of the CPLEX LP file format can be found here.

However, it is perhaps easiest to illustrate with some examples.
Consider \[\begin{array}{rrcrcrcrcl}
\min & x_1 & - & 2x_2 & +& x_3 &   &  &  \\
\mbox{s.t.}
 & x_1 & - &  2x_2 & + &  x_3 & - & x_4  & \geq & 1  \\
  & x_1 & + &  x_2 & + & 2x_3 & - & 2x_4 &  = & 3  \\
 &  & - &  x_2 & + & 3 x_3 & - & 5x_4  & \leq & 7  \\
   &  &  & & & x_3 & , & x_4 & \geq & 0.
\end{array}\]

The problem can be specified in LP format as follows:

\begin{verbatim}
min x_1 - 2x_2 + x_3
st
 x_1 - 2x_2 +  x_3 -  x_4 >= 1
 x_1 +  x_2 + 2x_3 - 2x_4  = 3
     -  x_2 + 3x_3 - 5x_4 <= 7
bounds
 x_1 free
 x_2 free
end
\end{verbatim}

Any variable that does not appear in the \texttt{bounds} section is
automatically assumed to be nonnegative.

\section{SoPlex}\label{soplex}

\href{http://soplex.zib.de/}{SoPlex} is an open-source linear
programming solver. It is free for noncommercial use.
Binaries for Mac OS X and Windows are readily available for
\href{http://soplex.zib.de/\#download}{download}.

One great feature of SoPlex is that it can return exact rational
solutions whereas most other solvers only return solutions as
floating-point numbers.

Suppose that the problem for the example at the beginning is saved in LP
format in an ASCII file named \texttt{eg.lp}. The following is the
output of running SoPlex in a macOS command-line terminal:

{\footnotesize
\begin{verbatim}
bash-3.2$ ./soplex-2.2.1.darwin.x86_64.gnu.opt -X --solvemode=2 -f=0 -o=0 eg.lp
SoPlex version 2.2.1 [mode: optimized] [precision: 8 byte] [rational: GMP 6.0.0] [githash: 267a44a]
Copyright (c) 1996-2016 Konrad-Zuse-Zentrum fuer Informationstechnik Berlin (ZIB)

int:solvemode = 2
real:feastol = 0
real:opttol = 0

Reading (real) LP file <eg.lp> . . .
Reading took 0.00 seconds.

LP has 3 rows 4 columns and 11 nonzeros.

Initial floating-point solve . . .
Simplifier removed 0 rows, 0 columns, 0 nonzeros, 0 col bounds, 0 row bounds
Reduced LP has 3 rows 4 columns 11 nonzeros
Equilibrium scaling LP

type |   time |   iters | facts |  shift   |violation |    value
  L  |    0.0 |       0 |     1 | 4.00e+00 | 2.00e+00 | 0.00000000e+00
  E  |    0.0 |       1 |     2 | 0.00e+00 | 4.00e+00 | 3.00000000e+00
  E  |    0.0 |       2 |     3 | 0.00e+00 | 0.00e+00 | 1.00000000e+00

Floating-point optimal.
Max. bound violation = 0
Max. row violation = 1/4503599627370496
Max. reduced cost violation = 0
Max. dual violation = 0
Performing rational reconstruction . . .
Tolerances reached.
Solved to optimality.

SoPlex status       : problem is solved [optimal]
Solving time (sec)  : 0.00
Iterations          : 2
Objective value     : 1.00000000e+00


Primal solution (name, value):
x_1               7/3
x_2               2/3
All other variables are zero.
bash-3.2$
\end{verbatim}
}

The option \texttt{-X} asks the solver to display the primal rational
solution. The options \texttt{-\/-solvemode=2} invokes iterative
refinement for solving for a rational solution. The options
\texttt{-f=0\ -o=0} set the primal feasibility and dual feasibility
tolerances to 0. Without these options, one might get only approximate
solutions to the problem. If we remove the last three options and
replace \texttt{-X} with \texttt{-x}, we obtain the following instead:

{\footnotesize
\begin{verbatim}
bash-3.2$ ./soplex-2.2.1.darwin.x86_64.gnu.opt -x eg.lp
SoPlex version 2.2.1 [mode: optimized] [precision: 8 byte] [rational: GMP 6.0.0] [githash: 267a44a]
Copyright (c) 1996-2016 Konrad-Zuse-Zentrum fuer Informationstechnik Berlin (ZIB)

Reading (real) LP file <eg.lp> . . .
Reading took 0.00 seconds.

LP has 3 rows 4 columns and 11 nonzeros.

Simplifier removed 0 rows, 0 columns, 0 nonzeros, 0 col bounds, 0 row bounds
Reduced LP has 3 rows 4 columns 11 nonzeros
Equilibrium scaling LP
type |   time |   iters | facts |  shift   |violation |    value
  L  |    0.0 |       0 |     1 | 4.00e+00 | 2.00e+00 | 0.00000000e+00
  E  |    0.0 |       1 |     2 | 0.00e+00 | 4.00e+00 | 3.00000000e+00
  E  |    0.0 |       2 |     3 | 0.00e+00 | 0.00e+00 | 1.00000000e+00
 --- transforming basis into original space
  L  |    0.0 |       0 |     1 | 0.00e+00 | 0.00e+00 | 1.00000000e+00
  L  |    0.0 |       0 |     1 | 0.00e+00 | 0.00e+00 | 1.00000000e+00

SoPlex status       : problem is solved [optimal]
Solving time (sec)  : 0.00
Iterations          : 2
Objective value     : 1.00000000e+00


Primal solution (name, value):
x_1       2.333333333
x_2       0.666666667
All other variables are zero (within 1.0e-16).
bash-3.2$ 
\end{verbatim}
}

There are many solver options that one can specify. To view the list of
all the options, simply run the solver without options and arguments.

\section{NEOS server for
optimization}\label{neos-server-for-optimization}

If one does not want to download and install SoPlex, one can use the
\href{https://neos-server.org/neos/solvers/index.html}{NEOS server for
optimization}. In addition to SoPlex, there are many other solvers to
choose from.

To solve a linear programming problem using SoPlex on the NEOS server,
one can submit a file in CPLEX LP format
\href{https://neos-server.org/neos/solvers/lp:SoPlex80bit/LP.html}{here}.

\section*{Exercises}\label{exercises-9}
\addcontentsline{toc}{section}{Exercises}

\begin{enumerate}
\def\labelenumi{\arabic{enumi}.}
\tightlist
\item
  Use SoPlex to obtain the exact optimal value of
  \[\begin{array}{rrcrcrcl}
  \min & 3 x & + & 2 y & + & 9 z \\
  \text{s.t.} & 53 x & + & 20 y & + & 96 z & \geq & 2 \\
   & 13 x & - & 7 y & + & 6 z & \geq & \frac{3}{17} \\
   & -x & + & 71 y & - & 3 z & \geq & \frac{7}{3} \\
   & x &, & y &,  &z  &\geq & 0.
  \end{array}\]
\end{enumerate}

\section*{Solutions}\label{solutions-9}
\addcontentsline{toc}{section}{Solutions}

\begin{enumerate}
\def\labelenumi{\arabic{enumi}.}
\tightlist
\item
  The optimal value is \(\frac{1882}{11679}\).
\end{enumerate}

