\section{Sets and Set Notation}

A set is a collection
\index{set notation} of things called elements.
For example $\left\{ 1,2,3,8\right\} $
would be a set consisting of the elements 1,2,3, and 8. To indicate that $3$
is an element of $\left\{ 1,2,3,8\right\} ,$ it is customary to write $3\in
\left\{ 1,2,3,8\right\} .$ We can also indicate when an element is not in a set, 
by writing $9\notin \left\{ 1,2,3,8\right\} $ which says that $9$ is
not an element of $\left\{ 1,2,3,8\right\} .$ Sometimes a rule specifies a
set. For example you could specify a set as all integers larger than $2.$
This would be written as $S=\left\{ x\in \mathbb{Z}:x>2\right\} .$ This
notation says: $S$ is the set of all integers, $x,$ such that $x>2.$

Suppose $A$ and $B$ are sets with the property that every element of $A$ is an
element of $B$. Then we say that $A$ is a subset of $B.$ For example, $\left\{
1,2,3,8\right\} $ is a subset of $\left\{ 1,2,3,4,5,8\right\}.$ In symbols, we write
$\left\{ 1,2,3,8\right\} \subseteq \left\{ 1,2,3,4,5,8\right\} .$ It is
sometimes said that ``$A$ is contained in $B$" or even ``$B$ contains $A$".
The same statement about the two sets may also be written as $\left\{
1,2,3,4,5,8\right\} \supseteq \left\{ 1,2,3,8\right\} $.

We can also talk about the {\em union\em} of two sets, which we write as $A \cup B$. This is the set consisting of everything which is an
element of at least one of the sets, $A$ or $B$. As an example of the union
of two sets, consider $\left\{ 1,2,3,8\right\} \cup \left\{ 3,4,7,8\right\} =\left\{
1,2,3,4,7,8\right\} $. This set is made up of the numbers which are in at least
one of the two sets.
\index{$\cup $}
\index{$\cap $} 

In general
\index{union}
\begin{equation*}
A\cup B = \left\{ x:x\in A
\text{ or }x\in B\right\} 
\end{equation*}
Notice that an element which is in {\em both\em} $A$ and $B$ is also in the
union, as well as elements which are in only one of $A$ or $B$. 

Another important set is the intersection \index{intersection} of two sets $A$ and $B$, written $A \cap B$. This set consists of everything which is in
{\em both\em} of the sets. Thus $\left\{ 1,2,3,8\right\} \cap \left\{ 3,4,7,8\right\}
=\left\{ 3,8\right\} $ because $3$ and $8$ are those elements the two sets
have in common. In general,
\begin{equation*}
A\cap B =  \left\{ x:x\in A\text{ and }x\in B\right\} 
\end{equation*}

If $A$ and $B$ are two sets, $A\setminus B$ \index{$\setminus$}denotes the set of things which
are in $A$ but not in $B.$ Thus
\begin{equation*}
A\setminus B =  \left\{ x\in A:x\notin B\right\} 
\end{equation*}
For example, if $A = \left\{1,2,3,8 \right\}$ and $B = \left\{ 3,4,7,8 \right\}$, then $A \setminus B = \left\{ 1,2,3,8\right\} \setminus 
\left\{ 3,4,7,8 \right\} =\left\{1,2 \right\}$.

A special set which is very important in mathematics is the empty set
\index{empty set} denoted by $\emptyset .$ The empty set, $\emptyset $, is
defined as the set which has no elements in it. It follows that the empty set is a subset of every set. 
This is true because if it were not so, there would have to exist a set $A,$ such that $\emptyset $
has something in it which is not in $A.$ However, $\emptyset $ has nothing
in it and so it must be that $\emptyset \subseteq A.$

We can also use brackets to denote sets which are intervals of numbers. Let $a$ and $b$ be real numbers. Then

\begin{itemize}
\item
$\left[ a,b\right] = \{x \in \mathbb{R}: a\leq x\leq b \}$

\item $\left[a,b \right) = \{x \in \mathbb{R}: a\leq x<b \}$ 

\item $\left( a,b\right) = \{x \in \mathbb{R}: a<x<b \}$ 

\item $(a,b] = \{ x \in \mathbb{R}: a<x\leq b \}$

\item  $[a,\infty ) = \{x \in \mathbb{R}: x\geq a \}$ 

\item $(-\infty ,a] = \{x \in \mathbb{R}: x \leq a \}$ 
\end{itemize}

These sorts of sets of real
numbers are called intervals. The two points $a$ and $b$ are called
endpoints, or bounds, of the interval. In particular, $a$ is the {\em lower bound \em}  while $b$ is the {\em upper bound \em} of the above
intervals, where applicable.
\index{intervals!notation} Other intervals such as $\left( -\infty ,b\right) $
are defined by analogy to what was just explained.
 In general, the curved
parenthesis, $($, indicates the end point is not included in the interval, while
the square parenthesis, $[$, indicates this end point is included. The reason that
there will always be a curved parenthesis next to $\infty $ or $-\infty $ is
that these are not real numbers and cannot be included in the interval in the way a real number can. 

To illustrate the use of this notation relative to intervals consider three
examples of inequalities. Their solutions will be written in the interval notation
just described.

\begin{example}{Solving an Inequality}{solvinginequality1}
Solve the inequality $2x+4\leq x-8$.
\end{example}

\begin{solution}
We need to find $x$ such that $2x+4\leq x-8$. Solving for $x$, we see that 
$x\leq -12$ is the answer. This is written in terms of an interval as $(-\infty ,-12].$
\end{solution}

Consider the following example.

\begin{example}{Solving an Inequality}{solvinginequality2}
Solve the inequality $\left( x+1\right) \left( 2x-3\right) \geq0.$
\end{example}

\begin{solution}
We need to find $x$ such that $\left( x+1\right) \left( 2x-3\right) \geq0.$ 
The solution is given by  $x\leq -1$ or $x\geq \frac{3}{2}$. Therefore, 
$x$ which fit into either of these intervals gives a solution. In terms of set notation this is denoted by $(-\infty ,-1]\cup
[ \vspace{0.05in}\frac{3}{2},\infty ).$
\end{solution}

Consider one last example.

\begin{example}{Solving an Inequality}{solvinginequality3}
Solve the inequality $x \left( x+2\right) \geq-4$.
\end{example}

\begin{solution}
This inequality is true for any value of $x$ where $x$ is a real number. We can write the solution as $\mathbb{R}$ or $\left(
-\infty ,\infty \right) .$
\end{solution}

In the next section, we examine another important mathematical concept.