\subsection{Cofactors and \texorpdfstring{$2\times 2$}{2x2} Determinants}

Let $A$ be an $n\times n$ matrix. That is, let $A$ be a square matrix. The \textbf{determinant} \index{determinant} of $A$, denoted
by $\det \left( A\right) $ is a very important number which we will explore throughout this section. 

If $A$ is a 2$\times 2$
matrix, the determinant is given by the following formula.

\begin{definition}{Determinant of a Two By Two Matrix}{twobytwodeterminant}
Let $A=\leftB
\begin{array}{rr}
a & b \\
c & d
\end{array}
\rightB .$ Then
\begin{equation*}
\det \left( A\right)  = ad-cb
\end{equation*}
\end{definition}

The determinant is also often denoted by enclosing the matrix with two
vertical lines. Thus
\begin{equation*}
\det \leftB
\begin{array}{rr}
a & b \\
c & d
\end{array}
\rightB =\left|
\begin{array}{rr}
a & b \\
c & d
\end{array}
\right| 
=ad - bc
\end{equation*}

The following is an example of finding the determinant of a $2 \times 2$ matrix.

\begin{example}{A Two by Two Determinant}{twobytwodeterminant}
Find $\det\left(A\right) $ for the matrix
$A =  \leftB
\begin{array}{rr}
2 & 4 \\
-1 & 6
\end{array}
\rightB .$
\end{example}

\begin{solution} From Definition \ref{def:twobytwodeterminant},
\begin{equation*}
\det \left( A\right) = \left( 2\right) \left( 6\right) -\left(
-1\right) \left( 4\right) = 12 + 4 = 16
\end{equation*}
\end{solution} 

The $2 \times 2$ determinant can be used to find the determinant of larger matrices.
We will now explore how to find the determinant of a $3 \times 3$ matrix, using several tools
including the $2 \times 2$ determinant.

We begin with the following definition. 

\begin{definition}{The $ij^{th}$ Minor of a Matrix}{ijthminor}
Let $A$ be a $3\times 3$ matrix. The $ij^{th}$ \textbf{minor} of $A$, denoted as
\index{determinant!minor} $minor\left( A\right) _{ij},$ is the determinant
of the $2\times 2$ matrix which results from deleting the $i^{th}$ row and
the $j^{th}$ column of $A$.

In general, if $A$ is an $n\times n$ matrix, then the $ij^{th}$ minor of $A$ is the determinant of the $n-1 \times n-1$ matrix which results from deleting the $i^{th}$ row and the $j^{th}$ column of $A$. 
\end{definition}

Hence, there is a minor associated with each entry of $A$. Consider the following example which demonstrates this definition. 

\begin{example}{Finding Minors of a Matrix}{findingminors}
Let 
\begin{equation*}
A = \leftB
\begin{array}{rrr}
1 & 2 & 3 \\
4 & 3 & 2 \\
3 & 2 & 1
\end{array}
\rightB 
\end{equation*}
Find $minor\left( A\right) _{12}$ and $minor\left( A\right) _{23}$.
\end{example}

\begin{solution} First we will find $minor\left( A\right) _{12}$. By Definition
\ref{def:ijthminor}, this is the determinant of the $2\times 2$ matrix
which results when you delete the first row and the second column. This
minor is given by
\begin{equation*}
minor \left(A\right)_{12}
=
\det \leftB
\begin{array}{rr}
4 & 2 \\
3 & 1
\end{array}
\rightB
\end{equation*}
Using Definition \ref{def:twobytwodeterminant}, we see that 
\begin{equation*}
\det \leftB
\begin{array}{rr}
4 & 2 \\
3 & 1
\end{array}
\rightB = \left(4\right)\left(1\right) - \left(3\right)\left(2\right) = 4 - 6 = -2
\end{equation*}

Therefore $minor \left(A\right)_{12} = -2$. 

Similarly, $minor\left(A\right)_{23}$ is the determinant of the $2\times 2$ matrix
which results when you delete the second row and the third column. This
minor is therefore
\begin{equation*}
minor \left(A\right)_{23} 
=
\det \leftB
\begin{array}{rr}
1 & 2 \\
3 & 2
\end{array}
\rightB = -4
\end{equation*}
Finding the other minors of $A$ is left as an exercise. 
\end{solution}

The $ij^{th}$ minor of a matrix $A$ is used in another important definition, given next.

\begin{definition}{The $ij^{th}$ Cofactor of a Matrix}{ijthcofactor}
Suppose $A$ is an $n\times n$ matrix. The $ij^{th}$ \textbf{cofactor}, denoted by $\func{cof}\left( A\right) _{ij}$ is
defined to be 
\begin{equation*}
\func{cof}\left( A\right) _{ij} = \left( -1\right) ^{i+j} minor\left(A\right)_{ij} 
\end{equation*}
\index{determinant!cofactor}
\end{definition}

It is also convenient to refer to the
cofactor of an entry of a matrix as follows. If $a_{ij}$ is the $ij^{th}$ entry of the
matrix, then its cofactor is just $\func{cof}\left( A\right) _{ij}.$

\begin{example}{Finding Cofactors of a Matrix}{findingcofactors}
Consider the matrix
\begin{equation*}
A=\leftB
\begin{array}{rrr}
1 & 2 & 3 \\
4 & 3 & 2 \\
3 & 2 & 1
\end{array}
\rightB 
\end{equation*}
Find $\func{cof}\left( A\right) _{12}$ and $\func{cof}\left( A\right) _{23}$.
\end{example}

\begin{solution} We will use Definition \ref{def:ijthcofactor} to compute these cofactors. 

First, we will compute $\func{cof}\left( A\right) _{12}$. 
Therefore, we need to find $minor\left(A\right)_{12}$. This is the determinant of the $2\times 2$ matrix
which results when you delete the first row and the second column. Thus $minor\left(A\right)_{12}$ is given by 
\begin{equation*}
\det \leftB
\begin{array}{rr}
4 & 2 \\
3 & 1
\end{array}
\rightB = -2
\end{equation*}
Then,
\begin{equation*}
\func{cof}\left( A\right) _{12}=\left( -1\right) ^{1+2} minor\left(A\right)_{12} =\left( -1\right) ^{1+2}\left( -2\right) =2
\end{equation*}
Hence, $\func{cof}\left( A\right) _{12}=2$.

Similarly, we can find $\func{cof}\left( A\right) _{23}$. First, find $minor\left(A\right)_{23}$, which is the determinant of the $2\times 2$ matrix
which results when you delete the second row and the third column. This
minor is therefore
\begin{equation*}
\det \leftB
\begin{array}{rr}
1 & 2 \\
3 & 2
\end{array}
\rightB = -4
\end{equation*}
Hence,
\begin{equation*}
\func{cof}\left( A\right) _{23}=\left( -1\right) ^{2+3} minor\left(A\right)_{23} =\left( -1\right) ^{2+3}\left( -4\right) =4
\end{equation*}
\end{solution}

You may wish to find the remaining cofactors for the above matrix. Remember that there is a cofactor 
for every entry in the matrix.  

We have now established the tools we need to find the determinant of a $3 \times3 $ matrix.

\begin{definition}{The Determinant of a Three By Three Matrix}{threebythreedeterminant}
Let $A$ be a $3\times 3$ matrix. Then, $\det \left(A\right)$ is calculated by picking a row
(or column) and taking the product of each entry in that row (column) with its
cofactor and adding these products together. 

This process when applied to the $i^{th}$
row (column) is known as \textbf{expanding along the $i^{th}$ row
(column)} \index{determinant!expanding along row or column} as is given by 
\[
\det \left(A\right) = a_{i1}\func{cof}(A)_{i1} + a_{i2}\func{cof}(A)_{i2} + a_{i3}\func{cof}(A)_{i3}
\]

\end{definition}

When calculating the determinant, you can choose to expand any row or any column. Regardless 
of your choice, you will always get the same number which is the determinant of the matrix $A.$ 
This method of evaluating a determinant by expanding along a row or a column is called \textbf{Laplace
Expansion} or \textbf{Cofactor Expansion}.
\index{Laplace expansion} \index{Cofactor Expansion}

Consider the following example. 

\begin{example}{Finding the Determinant of a Three by Three Matrix}{determinantthreebythree}
Let 
\begin{equation*}
A=\leftB
\begin{array}{rrr}
1 & 2 & 3 \\
4 & 3 & 2 \\
3 & 2 & 1
\end{array}
\rightB 
\end{equation*}
Find $\det\left(A\right)$ using the method of Laplace Expansion.
\end{example}

\begin{solution} First, we will calculate $\det \left(A\right)$ by expanding along the first column.
Using Definition \ref{def:threebythreedeterminant}, we take the $1$ in the first column and multiply it 
by its cofactor,
\begin{equation*}
1  \left( -1\right) ^{1+1}\left|
\begin{array}{rr}
3 & 2 \\
2 & 1
\end{array}
\right| 
=
(1)(1)(-1) = -1 
\end{equation*}
Similarly, we take the $4$ in the first column and multiply it by its cofactor, 
as well as with the $3$ in the first column. Finally, we add these numbers together, as given in the following
equation.
\begin{equation*}
\det \left(A\right) = 
1 \overset{
\func{cof}\left( A\right) _{11}}{\overbrace{\left( -1\right) ^{1+1}\left|
\begin{array}{rr}
3 & 2 \\
2 & 1
\end{array}
\right| }}+4 \overset{\func{cof}\left( A\right) _{21}}{\overbrace{\left(
-1\right) ^{2+1}\left|
\begin{array}{rr}
2 & 3 \\
2 & 1
\end{array}
\right| }}+3 \overset{\func{cof}\left( A\right) _{31}}{\overbrace{\left(
-1\right) ^{3+1}\left|
\begin{array}{rr}
2 & 3 \\
3 & 2
\end{array}
\right| }}
\end{equation*}
Calculating each of these, we obtain
\begin{equation*}
\det \left(A\right)
=
1 \left(1\right)\left(-1\right)
+
4 \left(-1\right)\left(-4\right)
+
3 \left(1\right)\left(-5\right)
=
-1 + 16 + -15
= 
0
\end{equation*}
Hence, $\det\left(A\right) = 0$. 

As mentioned in Definition \ref{def:threebythreedeterminant}, we can choose to 
expand along any row or column. Let's try now by expanding along the second row.
Here, we take the $4$ in the second row and multiply it to its cofactor, then 
add this to the $3$ in the second row multiplied by its cofactor, and the $2$ in the 
second row multiplied by its cofactor. The calculation is as follows. 
\begin{equation*}
\det \left(A\right)
=
4 \overset{\func{cof}\left( A\right) _{21}}{\overbrace{\left( -1\right)
^{2+1}\left|
\begin{array}{rr}
2 & 3 \\
2 & 1
\end{array}
\right| }}+3 \overset{\func{cof}\left( A\right) _{22}}{\overbrace{\left(
-1\right) ^{2+2}\left|
\begin{array}{rr}
1 & 3 \\
3 & 1
\end{array}
\right| }}+2 \overset{\func{cof}\left( A\right) _{23}}{\overbrace{\left(
-1\right) ^{2+3}\left|
\begin{array}{rr}
1 & 2 \\
3 & 2
\end{array}
\right| }}
\end{equation*}

Calculating each of these products, we obtain
\begin{equation*}
\det \left(A\right)
=
4\left(-1\right)\left(-2\right) 
+
3\left(1\right)\left(-8\right) 
+
2 \left(-1\right)\left(-4\right)
=
0
\end{equation*}

You can see that for both methods, we obtained $\det \left(A\right) = 0$. 
\end{solution}

As mentioned above, we will always come up with the same value for $\det \left(A\right)$ regardless of the row or column 
we choose to expand along. You should try to compute the above determinant by expanding along other
rows and columns. This is a good way to check your work, because you should come up 
with the same number each time!  

We present this idea formally in the following theorem.

\begin{theorem}{The Determinant is Well Defined}{welldefineddeterminant}
Expanding the $n\times n$ matrix along any row or column always gives the
same answer, which is the determinant.
\end{theorem}

We have now looked at the determinant of $2 \times 2$ and $3 \times 3$ matrices. It turns
out that the method used to calculate the determinant of a $3 \times 3$ matrix 
can be used to calculate the determinant of any sized matrix. Notice that
Definition \ref{def:ijthminor}, Definition \ref{def:ijthcofactor} and Definition \ref{def:threebythreedeterminant}
can all be applied to a matrix of any size. 

For example, the $ij^{th}$ minor of a $4 \times 4$ matrix is the determinant
of the $3 \times 3$ matrix you obtain when you delete the $i^{th}$ row and the $j^{th}$ column. 
Just as with the $3 \times 3$ determinant, we can compute the determinant of a $4 \times 4$ matrix by Laplace Expansion, 
along any row or column

Consider the following example. 

\begin{example}{Determinant of a Four by Four Matrix}{fourbyfourdeterminant}
Find $\det \left( A\right) $ where
\begin{equation*}
A=\leftB
\begin{array}{rrrr}
1 & 2 & 3 & 4 \\
5 & 4 & 2 & 3 \\
1 & 3 & 4 & 5 \\
3 & 4 & 3 & 2
\end{array}
\rightB
\end{equation*}
\end{example}

\begin{solution} As in the case of a $3\times 3$ matrix, you can expand this along any row or
column. Lets pick the third column. Then, using Laplace Expansion, 
\begin{equation*}
\det \left( A\right) 
=
3\left( -1\right) ^{1+3}\left\vert
\begin{array}{rrr}
5 & 4 & 3 \\
1 & 3 & 5 \\
3 & 4 & 2
\end{array}
\right\vert +2\left( -1\right) ^{2+3}\left\vert
\begin{array}{rrr}
1 & 2 & 4 \\
1 & 3 & 5 \\
3 & 4 & 2
\end{array}
\right\vert +
\end{equation*}
\begin{equation*}
4\left( -1\right) ^{3+3}\left\vert
\begin{array}{rrr}
1 & 2 & 4 \\
5 & 4 & 3 \\
3 & 4 & 2
\end{array}
\right\vert +3\left( -1\right) ^{4+3}\left\vert
\begin{array}{rrr}
1 & 2 & 4 \\
5 & 4 & 3 \\
1 & 3 & 5
\end{array}
\right\vert 
\end{equation*}

Now, you can calculate each $3 \times 3$ determinant using Laplace Expansion, as we did above. 
You should complete these as an exercise and verify that $\det \left( A \right)= -12$. 
\end{solution}

The following provides a formal definition for the determinant of an $n \times n$ matrix. You may
wish to take a moment and consider the above definitions for $2 \times 2$ and $3 \times 3$ determinants 
in context of this definition.

\begin{definition}{The Determinant of an $n\times n$ Matrix}{nbyndeterminant}
Let $A$ be an $n\times n$ matrix where $n\geq 2$ and suppose the
determinant of an $\left( n-1\right) \times \left( n-1\right) $ has been
defined. Then
\begin{equation*}
\det \left( A\right) =\sum_{j=1}^{n}a_{ij}\func{cof}\left( A\right)
_{ij}=\sum_{i=1}^{n}a_{ij}\func{cof}\left( A\right) _{ij}
\end{equation*}
The first formula consists of expanding the determinant along the $i^{th}$
row and the second expands the determinant along the $j^{th}$ column.
\end{definition}

In the following sections, we will explore some important properties and characteristics of the determinant.