\section{The General Solution of a Linear System}

\begin{outcome}
\begin{enumerate}
\item[A.]  Use linear transformations to determine the particular solution and general solution to a system of equations.

\item[B.]  Find the kernel of a linear transformation. 
\end{enumerate}
\end{outcome}

Recall the definition of a linear transformation discussed above. 
$T$ is a \textbf{linear transformation} if whenever $\vect{x}, \vect{y}$ are
vectors and $k,p$ are scalars,
\begin{equation*}
T\left( k\vect{x}+p\vect{y}\right) =k T \left( \vect{x} \right) +p T\left(\vect{y} \right)
\end{equation*}
Thus linear transformations distribute across addition and pass scalars to
the outside.

It turns out that we can use linear transformations to solve linear
systems of equations. Indeed given a system of linear equations of the
form $A\vect{x}=\vect{b}$, one may rephrase this as $T(\vect{x})=\vect{b}$ where $T$ is the linear
transformation $T_A$ induced by the coefficient matrix $A$. With this in mind consider the following definition. 

\begin{definition}{Particular Solution of a System of Equations}{particularsolutions}
Suppose a linear system of equations can be written in the form
\begin{equation*}
T\left(\vect{x}\right)=\vect{b}
\end{equation*}
If $T\left(\vect{x}_{p}\right)=\vect{b},$ 
then $\vect{x}_{p}$ is called a \textbf{particular solution} of
\index{particular solution} the linear system.
\end{definition}

Recall that a system is called homogeneous if every equation in the system is equal to $0$. 
Suppose we represent a homogeneous system of equations by $T\left(\vect{x}\right)=0$. It turns out
that the $\vect{x}$ for which $T \left(\vect{x}\right) = 0$ are part of a special set called the \textbf{null space}
of $T$. We may also refer to the null space as the \textbf{kernel} of $T$, and we write $ker\left(T\right)$. 

Consider the following definition.

\begin{definition}{Null Space or Kernel of a Linear Transformation}{nullspaceoflineartransformation}
Let $T$ be a linear transformation. Define
\begin{equation*}
\ker \left( T\right) = \left\{ \vect{x}:T \left(\vect{x} \right)= \vect{0} \right\} 
\end{equation*}
The kernel, \index{kernel} $\ker \left( T\right) $ consists of the set of all vectors $\vect{x}$ for which
$T (\vect{x}) = \vect{0}$. This is also called the
\index{null space} \textbf{null space} of $T$. 
\end{definition}

We may also refer to the kernel of $T$ as the
\textbf{solution space }of the equation $T \left(\vect{x}\right) = \vect{0}$.
\index{solution space}

Consider the following example.

\begin{example}{The Kernel of the Derivative}{kernelderivative}
Let $\frac{d}{dx}$ denote the linear transformation defined on $f,$ the functions
which are defined on $\mathbb{R}$ and have a continuous derivative. Find 
$\ker \left( \frac{d}{dx}\right) .$
\end{example}

\begin{solution} The example asks for functions $f$ which the property that $\frac{df}{dx}
=0. $ As you may know from calculus, these functions are the constant functions.
Thus $\ker \left( \frac{d}{dx}\right)$ is the set of constant functions.
\end{solution} 

Definition \ref{def:nullspaceoflineartransformation} states that $\ker \left( T\right) $ is the set of
solutions to the equation,
\begin{equation*}
T\left( \vect{x} \right) = \vect{0}
\end{equation*}
Since we can write $T\left( \vect{x} \right)$ as $A\vect{x}$, you have been solving such
equations for quite some time.

We have spent a lot of time finding solutions to systems of equations in general, as well as
homogeneous systems. Suppose we look at a system given by $A\vect{x}=\vect{b}$, and consider the 
related homogeneous system. By this, we mean that we replace $\vect{b}$ by $\vect{0}$ and look at $A\vect{x}=\vect{0}$. 
It turns out that there is a very important relationship between the solutions of the original
system and the solutions of the associated homogeneous system. In the following 
theorem, we use linear transformations to denote a system of equations. Remember that
$T\left(\vect{x}\right) = A\vect{x}$.

\begin{theorem}{Particular Solution and General Solution}{particularandgeneralsolution}
Suppose $\vect{x}_{p}$ is a solution to the linear system given by ,
\begin{equation*}
T\left( \vect{x} \right) = \vect{b}
\end{equation*}
Then if $\vect{y}$ is any other solution to $T\left(\vect{x}\right)=\vect{b}$, 
\index{general solution!solution space} there exists $\vect{x}_0 \in \ker
\left( T\right) $ such that
\begin{equation*}
\vect{y} = \vect{x}_{p}+ \vect{x}_0
\end{equation*}
Hence, every solution to the linear system can be written as a sum of a particular solution, $\vect{x}_p$,
 and a solution $\vect{x}_0$ to the associated 
homogeneous system given by $T\left(\vect{x}\right)=\vect{0}$.
\end{theorem}

\begin{proof}
Consider $\vect{y} - \vect{x}_{p}= \vect{y} + \left(
-1\right) \vect{x}_{p}.$ Then $T\left( \vect{y} - \vect{x}_{p}\right) =T\left(\vect{y}\right)
-T\left( \vect{x}_{p} \right)$. Since $\vect{y}$ and $\vect{x}_{p}$ are both solutions to the system, it follows that $T\left(\vect{y}\right)= \vect{b} $
and $T\left(\vect{x}_p\right) = \vect{b}$. 

Hence, $T\left(\vect{y}\right)-T\left( \vect{x}_{p} \right)
=\vect{b} - \vect{b} = \vect{0}$.  Let $\vect{x}_0 = \vect{y} - \vect{x}_{p}$.
Then, $T\left(\vect{x}_0\right)= \vect{0} $ so $\vect{x}_0$ is a solution to the associated homogeneous system and so is in $\ker \left(T\right)$.
\end{proof}

Sometimes people remember the above theorem in the following form. The
solutions to the system $T\left(\vect{x}\right)=\vect{b}$ are given by 
$\vect{x}_{p}+\ker \left( T\right) $ where $\vect{x}_{p}$ is a particular
solution to $T\left(\vect{x}\right)=\vect{b}$.

For now, we have been speaking about the kernel or null space of a linear transformation $T$. However, 
we know that every linear transformation $T$ is determined by some matrix $A$. Therefore,
we can also speak about the null space of a matrix. Consider the following example.  

\begin{example}{The Null Space of a Matrix}{matrixnullspace}
Let
\begin{equation*}
A=\leftB
\begin{array}{rrrr}
1 & 2 & 3 & 0 \\
2 & 1 & 1 & 2 \\
4 & 5 & 7 & 2
\end{array}
\rightB
\end{equation*}
Find $\func{null} \left( A\right) $. Equivalently, find the solutions to the
system of equations $A\vect{x}=\vect{0}$.
\end{example}

\begin{solution}  We are asked to find $\left\{ \vect{x} : A\vect{x} = \vect{0}\right\} .$ In other
words we want to solve the system, $A\vect{x}=\vect{0}$. Let $\vect{x} =
\leftB
\begin{array}{r}
x \\
y \\
z \\
w
\end{array}
\rightB.$ Then this amounts to solving
\begin{equation*}
\leftB
\begin{array}{rrrr}
1 & 2 & 3 & 0 \\
2 & 1 & 1 & 2 \\
4 & 5 & 7 & 2
\end{array}
\rightB \leftB
\begin{array}{c}
x \\
y \\
z \\
w
\end{array}
\rightB =\leftB
\begin{array}{c}
0 \\
0 \\
0
\end{array}
\rightB
\end{equation*}
This is the linear system
\begin{equation*}
\begin{array}{c}
x+2y+3z=0 \\
2x+y+z+2w=0 \\
4x+5y+7z+2w=0
\end{array}
\end{equation*}
To solve, set up the augmented matrix and row reduce to find the \rref.
\begin{equation*}
\leftB
\begin{array}{rrrr|r}
1 & 2 & 3 & 0 & 0 \\
2 & 1 & 1 & 2 & 0 \\
4 & 5 & 7 & 2 & 0
\end{array}
\rightB
\rightarrow \cdots \rightarrow
\leftB
\begin{array}{rrrr|r}
1 & 0 & -
\vspace{0.05in}\frac{1}{3} & \vspace{0.05in}\frac{4}{3} &  0 \\
0 & 1 & \vspace{0.05in}\frac{5}{3} & -\vspace{0.05in}\frac{2}{3} & 0 \\
0 & 0 & 0 & 0 & 0
\end{array}
\rightB 
\end{equation*}
This yields $x=\vspace{0.05in}\frac{1}{3}z-\vspace{0.05in}\frac{4}{3}w$ and 
$y=\vspace{0.05in}\frac{2}{3}w-\vspace{0.05in}\frac{5}{3}z.$
Since $\func{null} \left( A\right) $ consists of the solutions to this system, it consists vectors of the form,
\begin{equation*}
\leftB
\begin{array}{c}
\vspace{0.05in}\frac{1}{3}z-\vspace{0.05in}\frac{4}{3}w \\
\vspace{0.05in}\frac{2}{3}w-\vspace{0.05in}\frac{5}{3}z \\
z \\
w
\end{array}
\rightB =z \leftB
\begin{array}{r}
\vspace{0.05in}\frac{1}{3} \\
-\vspace{0.05in}\frac{5}{3} \\
1 \\
0
\end{array}
\rightB +w \leftB
\begin{array}{r}
-\vspace{0.05in}\frac{4}{3} \\
\vspace{0.05in}\frac{2}{3} \\
0 \\
1
\end{array}
\rightB 
\end{equation*}
\end{solution}

Consider the following example.

\begin{example}{A General Solution}{generalsolution}
The \textbf{general solution} \index{general solution} of a linear system of equations 
is the set of
all possible solutions. Find the general solution to the linear system,
\begin{equation*}
\leftB
\begin{array}{rrrr}
1 & 2 & 3 & 0 \\
2 & 1 & 1 & 2 \\
4 & 5 & 7 & 2
\end{array}
\rightB \leftB
\begin{array}{r}
x \\
y \\
z \\
w
\end{array}
\rightB =\leftB
\begin{array}{r}
9 \\
7 \\
25
\end{array}
\rightB
\end{equation*}
given that $\leftB
\begin{array}{r}
x \\
y \\
z \\
w
\end{array}
\rightB=\leftB
\begin{array}{r}
1 \\
1 \\
2 \\
1
\end{array}
\rightB$ is one solution.
\end{example}

\begin{solution} Note the matrix of this system is the same as the matrix in Example 
\ref{exa:matrixnullspace}. Therefore, from Theorem \ref{thm:particularandgeneralsolution}, you will obtain all
solutions to the above linear system by adding a particular solution $\vect{x}_p$ to the solutions of the associated homogeneous 
system, $\vect{x}$. One particular solution is given above by
\begin{equation}
\vect{x}_p
=
\leftB
\begin{array}{r}
x \\
y \\
z \\
w
\end{array}
\rightB=\leftB
\begin{array}{r}
1 \\
1 \\
2 \\
1
\end{array}
\rightB
\end{equation}

Using this particular solution along with the solutions found in Example \ref{exa:matrixnullspace}, we
obtain the following solutions, 
\begin{equation*}
z\leftB
\begin{array}{r}
\vspace{0.05in}\frac{1}{3} \\
-\vspace{0.05in}\frac{5}{3} \\
1 \\
0
\end{array}
\rightB +w\leftB
\begin{array}{r}
-\vspace{0.05in}\frac{4}{3} \\
\vspace{0.05in}\frac{2}{3} \\
0 \\
1
\end{array}
\rightB +\leftB
\begin{array}{r}
1 \\
1 \\
2 \\
1
\end{array}
\rightB 
\end{equation*}

Hence, any solution to the above linear system is of this form.
\end{solution} 