\section{Vectors in $\mathbb{R}^n$}

\begin{outcome}

\begin{enumerate}
\item[A.] Find the position vector of a point in $\mathbb{R}^n$.
\end{enumerate}
\end{outcome}

The notation $\mathbb{R}^{n}$ refers to the collection of ordered lists of 
$n$ real numbers, that is 
\[
\mathbb{R}^{n} = \left\{ \left( x_{1}\cdots x_{n}\right)
:x_{j}\in \mathbb{R}\text{ for }j=1,\cdots ,n\right\}
\]
In this chapter, we take a closer look at vectors in $\mathbb{R}^n$. First, we will consider what $\mathbb{R}^n$ looks like 
in more detail. Recall that the point given by $0=\left( 0, \cdots, 0 \right)$ is called the \textbf{origin}.

Now, consider the case of $\mathbb{R}^n$ for $n=1.$ Then from the definition we can identify 
$\mathbb{R}$ with points in $\mathbb{R}^{1}$ as follows:
\begin{equation*}
\mathbb{R} = \mathbb{R}^{1}=
 \left\{ \left( x_{1}\right) :x_{1}\in \mathbb{R} \right\}
\end{equation*}
Hence, $\mathbb{R}$ is defined as the set of all real numbers and geometrically,
we can describe this as all the points on a line.

Now suppose $n=2$. Then, from the definition,
\begin{equation*}
\mathbb{R}^{2}=
\left\{ \left(x_{1}, x_{2}\right)
:x_{j}\in \mathbb{R}\text{ for }j=1,2 \right\}
\end{equation*}
Consider the
familiar coordinate plane, with an $x$ axis and a $y$ axis. Any point
within this coordinate plane is identified by where it is located
along the $x$ axis, and also where it is located along the $y$
axis. Consider as an example the following diagram.

\begin{center}
\begin{tikzpicture}
\draw[<->](-4,0)--(4,0);
\draw[<->](0,-1)--(0,5);
\draw(2,0.2)--(2,-0.2);
\draw(-3,0.2)--(-3,-0.2);
\draw(-0.2,1)--(0.2,1);
\draw(-0.2,4)--(0.2,4);
\draw[help lines] (-3,0)--(-3,4)--(0,4);
\draw[help lines](0,1)--(2,1)--(2,0);
\draw[fill=red](-3,4) circle [radius=3pt];
\draw[fill=red](2,1) circle [radius=3pt];
\node[below right] at (4,0){$x$};
\node[left] at(0,5){$y$};
\node[below] at (-3,-0.5){$-3$};
\node[below] at (2,-0.5){$2$};
\node[left] at (-0.5, 1){$1$};
\node[left] at (-0.5, 4){$4$};
\node[above] at (-3,4){$Q = (-3,4)$};
\node[above] at (2,1){$P = (2,1)$};
\end{tikzpicture}
\end{center}

Hence, every element in $\mathbb{R}^2$ is identified by two
components, $x$ and $y$, in the usual manner. The coordinates $x, y$ (or $x_1$,$x_2$) uniquely determine a point in the plan. Note that while the definition uses $x_1$ and $x_2$ to label the coordinates and you may be used to $x$ and $y$, these notations are equivalent. 

Now suppose $n=3$. You may have previously encountered the $3$-dimensional
coordinate system, given by
\begin{equation*}
\mathbb{R}^{3}=
\left\{ \left( x_{1}, x_{2}, x_{3}\right)
:x_{j}\in \mathbb{R}\text{ for }j=1,2,3 \right\}
\end{equation*}

Points in $\mathbb{R}^3$ will be determined by three
coordinates, often written $\left(x,y,z\right)$ which correspond to the $x$, $y$,
and $z$ axes.  We can think as above that the first two coordinates
determine a point in a plane. The third component determines the
height above or below the plane, depending on whether this number is
positive or negative, and all together this determines a point in
space. You
see that the ordered triples correspond to points in space just as the
ordered pairs correspond to points in a plane and single real numbers
correspond to points on a line.

The idea behind the more general $\mathbb{R}^n$ is that we can extend
these ideas beyond $n = 3.$ This discussion regarding points in $\mathbb{R}^n$ leads into a study of vectors in $\mathbb{R}^n$. While we consider $\mathbb{R}^n$ for all $n$,
we will largely  focus on $n=2,3$ in this section. 

Consider the following definition.

\begin{definition}{The Position Vector}{positionvector}
Let $P=\left( p_{1},\cdots ,p_{n}\right) $ be the coordinates of a
point in $\mathbb{R}^{n}.$ Then the vector $\longvect{0P}$ with its tail at
$0=\left( 0,\cdots ,0\right) $ and its tip at 
$P$ is called the \textbf{position vector }of
\index{position vector} \index{vector!points and vectors} the point $P$. 
We write 
\begin{equation*}
\longvect{0P} = \leftB 
\begin{array}{c}
p_{1} \\
\vdots \\
p_{n}
\end{array} \rightB
\end{equation*}
\end{definition}

For this reason we may write both $P=\left( p_{1},\cdots ,p_{n}\right)  \in \mathbb{R}^{n}$
and  $\longvect{0P} = \leftB p_{1} \cdots p_{n} \rightB^T \in \mathbb{R}^{n}$. 

This definition is illustrated in the following picture for the
special case of $\mathbb{R}^{3}$.

\begin{center}
\begin{tikzpicture}
\draw[<->](2,0,0)--(0,0,0)--(0,2,0);
\draw[->](0,0,0)--(0,0,2);
\draw[thick, blue](0,0,0)--(1.5,1,0);
\draw[fill=red] (1.5, 1,0) circle [radius=3pt];
\node[above right] at (1.5, 1.25, 0){$P=(p_1, p_2, p_3)$};
\node[below right] at (1.5, 0.75, 0){$\longvect{0P}= \leftB \begin{array}{ccc}
p_1 & p_2 & p_3 
\end{array}\rightB^T$};
\end{tikzpicture}
\end{center}

Thus every point $P$ in $\mathbb{R}^{n}$ determines its position
vector $\longvect{0P}$. Conversely, every such position vector $\longvect{0P}$
which has its tail at $0$ and point at $P$ determines the point $P$ of
$\mathbb{R}^{n}$.

Now suppose we are given two points, $P,Q$ whose
coordinates are $\left( p_{1},\cdots ,p_{n}\right) $ and $\left(
q_{1},\cdots ,q_{n}\right) $ respectively. We can also determine the
\textbf{position vector from $P$ to $Q$} (also called the \textbf{vector from $P$ to $Q$}) defined as follows.
\begin{equation*}
\longvect{PQ} = \leftB \begin{array}{c}
q_{1}-p_{1}\\
 \vdots \\
q_{n}-p_{n}
\end{array}
\rightB = \longvect{0Q} - \longvect{0P}
\end{equation*}

Now, imagine taking a vector in $\mathbb{R}^n$ and moving it around, always
keeping it pointing in the same direction as shown in the following picture.

\begin{center}
\begin{tikzpicture}
\draw[<->](2,0,0)--(0,0,0)--(0,2,0);
\draw[->](0,0,0)--(0,0,2);
\draw[->, thick, blue](0,0,0)--(1.5,1,0);
\draw[->, thick, blue](-1,1,0)--(0.5,2,0);
\node[right] at (1.5, 1.25, 0){$\longvect{0P}= \leftB \begin{array}{ccc}
p_1 & p_2 & p_3 \end{array}\rightB ^T$};
\node[below] at (-1,1,0){$A$};
\node[below] at (0.5,2,0){$B$};
\end{tikzpicture}
\end{center}

After moving it around, it is regarded as the same vector. Each vector, $\longvect{0P}$ and $\longvect{AB}$ has the same length (or magnitude) and direction. Therefore, they are equal. 

Consider now the general definition for a vector in $\mathbb{R}^n$. 

\begin{definition}{Vectors in $\mathbb{R}^n$}{vectorsRn}
Let $\mathbb{R}^{n} = \left\{ \left( x_{1}, \cdots, x_{n}\right)
:x_{j}\in \mathbb{R}\text{ for }j=1,\cdots ,n\right\} .$
Then,
\begin{equation*}
\vect{x}
=
\leftB \begin{array}{c}
x_{1} \\
\vdots \\
x_{n}
\end{array}
\rightB 
\end{equation*}
is called a \textbf{vector}. Vectors have both size (magnitude) and direction. 
The numbers $x_{j}$ are called the \textbf{components}  of $\vect{x}$. 
\index{vector!components}
\index{vector}
\end{definition}

Using this notation, we may use $\vect{p}$ to denote the position vector of point $P$. Notice that in this context, $\vect{p} = \longvect{0P}$. These notations may be used interchangeably. 

You can think of the components of a vector as directions for
obtaining the vector. Consider $n=3$.  Draw a vector with its tail at
the point $\left( 0,0,0\right) $ and its tip at the point $\left(
a,b,c\right) $. This vector it is obtained by starting at $\left(
0,0,0\right) $, moving parallel to the $x$ axis to $\left(
a,0,0\right) $ and then from here, moving parallel to the $y$ axis to
$\left( a,b,0\right) $ and finally parallel to the $z$ axis to $\left(
a,b,c\right). $ Observe that the same vector would result if you began
at the point $ \left( d,e,f \right)$, moved parallel to the $x$ axis
to $\left( d+a,e,f\right) ,$ then parallel to the $y$ axis to $\left(
d+a,e+b,f\right) ,$ and finally parallel to the $z$ axis to $\left(
d+a,e+b,f+c\right)$. Here, the vector would have its tail sitting at
the point determined by $A= \left( d,e,f\right) $ and its point at
$B=\left( d+a,e+b,f+c\right) .$ It is the \textbf{same vector} because
it will point in the same direction and have the same length. It is
like you took an actual arrow, \index{vector!addition, geometric
meaning} and moved it from one location to another keeping it pointing
the same direction.

We conclude this section with a brief discussion regarding notation. In previous sections, we have written vectors as columns, or $n \times 1$ matrices. For convenience in this chapter we may write vectors as the transpose of row vectors, or $1 \times n$ matrices. These are of course equivalent and we may move between both notations. Therefore, recognize that 
\[
\leftB
\begin{array}{r}
2 \\
3
\end{array}
\rightB  = \leftB
\begin{array}{rr}
2 & 3
\end{array}
\rightB^T 
\]

Notice that two vectors $\vect{u} = \leftB u_{1} \cdots u_{n}\rightB^T $ and
$\vect{v}=\leftB v_{1} \cdots v_{n}\rightB^T$ are equal if and only if
all corresponding components are equal. Precisely,
\begin{equation*}
\begin{array}{c}
\vect{u}=\vect{v} \; \mbox{if and only if}\\
u_{j}=v_{j} \; \mbox{for all}\; j=1,\cdots ,n
\end{array}
\end{equation*} 
Thus 
$\leftB 
\begin{array}{rrr}
1  & 2 & 4
\end{array}
\rightB^T \in \mathbb{R}^{3}$ and $\leftB 
\begin{array}{rrr}
2 & 1 & 4
\end{array}
\rightB^T \in
\mathbb{R}^{3}$ but $\leftB 
\begin{array}{rrr}
1 & 2 & 4
\end{array}
\rightB^T \neq \leftB
\begin{array}{rrr}
2 & 1 & 4
\end{array}
\rightB^T $ because,
even though the same numbers are involved, the order of the numbers is different. 

For the specific case of $\mathbb{R}^3$, there are three special vectors which we often use. 
They are given by 
\begin{equation*}
\vect{i} = 
\leftB
\begin{array}{rrr}
1 & 0 & 0
\end{array}
\rightB^T
\end{equation*}
\begin{equation*}
\vect{j} = 
\leftB
\begin{array}{rrr}
0 & 1 & 0
\end{array}
\rightB^T
\end{equation*}
\begin{equation*}
\vect{k} = 
\leftB
\begin{array}{rrr}
0 & 0 & 1
\end{array}
\rightB^T
\end{equation*}
We can write any vector $\vect{u} = 
\leftB
\begin{array}{rrr}
u_1 & u_2 & u_3
\end{array}
\rightB^T$
as a linear combination of these vectors, written as $\vect{u} = u_1 \vect{i} + u_2 \vect{j} + u_3 \vect{k}$. This notation will be used throughout 
this chapter.