\subsection{Orthogonal and Orthonormal Sets}

In this section, we examine what it means for vectors (and sets of
vectors) to be orthogonal and orthonormal. First, it is necessary to
review some important concepts. You may recall the definitions for the
span of a set of vectors and a linear independent set of vectors. We
include the definitions and examples here for convenience.

\begin{definition}{Span of a Set of Vectors and Subspace}{span}
The collection of all linear combinations of a set of vectors $\{ \vec{u}_1,
\cdots ,\vec{u}_k\}$ in $\mathbb{R}^{n}$ is known as the span of these
vectors and is written as $\func{span} \{\vec{u}_1, \cdots ,
\vec{u}_k\}$. \\ 
We call a collection of the form $\func{span} \{\vec{u}_1, \cdots , \vec{u}_k\}$ 
a subspace of $\mathbb{R}^{n}$.
\index{span} \index{vectors!span}
\end{definition}

Consider the following example. 

\begin{example}{Span of Vectors}{spanvectors}
Describe the span of the vectors $\vec{u}=\leftB 
\begin{array}{rrr}
1  & 1 & 0
\end{array}
\rightB^T$ and
$\vec{v}=\leftB 
\begin{array}{rrr}
3  & 2 & 0
\end{array}
\rightB^T \in \mathbb{R}^{3}$.
\end{example}

\begin{solution}
You can see that any linear combination of the vectors $\vect{u}$ and $\vect{v}$ yields a vector 
$\leftB 
\begin{array}{rrr}
x  & y & 0
\end{array}
\rightB^T$ in the $XY$-plane. 

Moreover every vector in the $XY$-plane is in fact such a linear
combination of the vectors $\vect{u}$ and $\vect{v}$. That's because
\[ \leftB 
\begin{array}{r}
x \\
y \\
0
\end{array}
\rightB 
=
(-2x+3y) \leftB 
\begin{array}{r}
1 \\
1 \\
0
\end{array}
\rightB
+
(x-y)\leftB 
\begin{array}{r}
3 \\
2 \\
0
\end{array}
\rightB 
\]

Thus  span$\{\vect{u},\vect{v}\}$ is precisely the $XY$-plane.
\end{solution}

The span of a set of a vectors in $\mathbb{R}^n$ is what we call a \textbf{subspace of $\mathbb{R}^n$}. A subspace $W$ is characterized by the feature that any linear combination of vectors of $W$ is again a vector contained in $W$. 

Another important property of sets of vectors is called linear independence. 

\begin{definition}{Linearly Independent Set of Vectors}{linearindependentortho}
A set of non-zero vectors $\{ \vect{u}_1, \cdots ,\vect{u}_k\}$ in $\mathbb{R}^{n}$ is said to be 
\textbf{linearly independent} if no vector in that set is in the span of the other vectors of that set.
\index{linear independent}\index{vectors!linear independent}
\end{definition}

Here is an example.  

\begin{example}{Linearly Independent Vectors}{linearlyindependentvectors}
Consider vectors $\vect{u}=\leftB 
\begin{array}{rrr}
1  & 1 & 0
\end{array}
\rightB^T$, 
$\vect{v}=\leftB 
\begin{array}{rrr}
3  & 2 & 0
\end{array}
\rightB^T$, and
$\vect{w}=\leftB 
\begin{array}{rrr}
4  & 5 & 0
\end{array}
\rightB^T  \in \mathbb{R}^{3}$.
Verify whether the set $\{\vect{u}, \vect{v}, \vect{w}\}$ is linearly independent. 
\end{example}

\begin{solution}
We already verified in Example \ref{exa:spanvectors} that
$\func{span} \{\vect{u}, \vect{v} \}$ is the $XY$-plane. Since $\vect{w}$
is clearly also in the $XY$-plane, then the set $\{\vect{u}, \vect{v},
\vect{w}\}$ is \textbf{not} linearly independent.
\end{solution}

In terms of spanning, a set of vectors is linearly independent if it
does not contain unnecessary vectors. In the previous example you
can see that the vector $\vect{w}$ does not help to span any new vector
not already in the span of the other two vectors. However you can
verify that the set $\{\vect{u}, \vect{v}\}$ is linearly independent,
since you will not get the $XY$-plane as the span of a single vector.

We can also determine if a set of vectors is linearly independent by examining linear combinations. A set of vectors is linearly independent if and only if
whenever a linear combination of these vectors equals zero, it follows
that all the coefficients equal zero. It is a good exercise to verify
this equivalence, and this latter condition is often used as the
(equivalent) definition of linear independence.

If a subspace is spanned by a linearly independent set of vectors,
then we say that it is a basis for the subspace.

\begin{definition}{Basis of a Subspace}{subspacebasis}
Let $V$ be a subspace of $\mathbb{R}^{n}$. Then $\left\{
\vect{u}_{1},\cdots ,\vect{u}_{k}\right\} $ is a \textbf{basis} for
$V$ if the following two conditions
hold. \index{basis}\index{vectors!basis}\index{subspace!basis}

\begin{enumerate}
\item $\func{span}\left\{ \vect{u}_{1},\cdots ,\vect{u}_{k}\right\} =V$
\item $\left\{ \vect{u}_{1},\cdots ,\vect{u}_{k}\right\} $ is linearly
independent
\end{enumerate}
\end{definition}

Thus the set of vectors $\{\vect{u}, \vect{v}\}$ from Example
\ref{exa:linearlyindependentvectors} is a basis for $XY$-plane in
$\mathbb{R}^{3}$ since it is both linearly independent and spans
the $XY$-plane.

 Recall from the properties of the dot product of vectors 
that two vectors $\vect{u}$ and $\vect{v}$ are orthogonal if $\vect{u}
\dotprod \vect{v} = 0$. Suppose a vector is orthogonal to a spanning set of $\mathbb{R}^n$. What can be said about such a vector? This is the discussion in the following example.

\begin{example}{Orthogonal Vector to a Spanning Set}{}
Let $\{\vect{x}_1, \vect{x}_2, \ldots, \vect{x}_k\}\in\mathbb{R}^n$ and
suppose $\mathbb{R}^n=\func{span}\{\vect{x}_1, \vect{x}_2, \ldots, \vect{x}_k\}$.
Furthermore, suppose that there exists a vector $\vect{u}\in\mathbb{R}^n$ for which $\vect{u}\dotprod \vect{x}_j=0$ for all $j$, $1\leq j\leq k$.
What type of vector is $\vect{u}$?
\end{example}

\begin{solution}
Write $\vect{u}=t_1\vect{x}_1 + t_2\vect{x}_2 +\cdots +t_k\vect{x}_k$
for some $t_1, t_2, \ldots, t_k\in\mathbb{R}$
(this is possible because
$\vect{x}_1, \vect{x}_2, \ldots, \vect{x}_k$ span $\mathbb{R}^n$).
 
Then
\begin{eqnarray*}
\vectlength\vect{u}\vectlength^2 & = & \vect{u}\dotprod\vect{u} \\
& = & \vect{u}\dotprod(t_1\vect{x}_1 + t_2\vect{x}_2 +\cdots +t_k\vect{x}_k) \\
& = & \vect{u}\dotprod (t_1\vect{x}_1) +  \vect{u}\dotprod (t_2\vect{x}_2) +
\cdots +  \vect{u}\dotprod (t_k\vect{x}_k) \\
& = & t_1(\vect{u}\dotprod \vect{x}_1) + t_2(\vect{u}\dotprod \vect{x}_2) + \cdots 
+ t_k(\vect{u}\dotprod \vect{x}_k) \\
& = & t_1(0) + t_2(0) + \cdots + t_k(0) = 0.
\end{eqnarray*}
Since $\vectlength\vect{u}\vectlength^2 =0$, $\vectlength\vect{u}\vectlength =0$.
We know that $\vectlength\vect{u}\vectlength=0$ if and only if 
$\vect{u}=\vect{0}_n$.
Therefore, $\vect{u}=\vect{0}_n$.
In conclusion, the only vector orthogonal to every vector of
a spanning set of $\mathbb{R}^n$ is the zero vector.
\end{solution}

We can now discuss what is meant by an orthogonal set of vectors. \index{orthogonal}

\begin{definition}{Orthogonal Set of Vectors}{orthset}
Let $\{ \vect{u}_1, \vect{u}_2, \cdots, \vect{u}_m \}$ be a set of
vectors in $\mathbb{R}^n$. Then this set is called an
\textbf{orthogonal set} if the following conditions hold:
\index{orthogonal}\index{vectors!orthogonal}
\begin{enumerate}
\item 
$\vect{u}_i \dotprod \vect{u}_j = 0$ for all $i \neq j$ 
\item
$\vect{u}_i \neq \vect{0}$ for all $i$
\end{enumerate}
\end{definition}

If we have an orthogonal set of vectors and normalize each vector so
they have length 1, the resulting set is called an \textbf{orthonormal
set} of vectors. They can be described as follows.

\begin{definition}{Orthonormal Set of Vectors}{orthoSetVectors}
A set of vectors, $\left\{ \vect{w}_{1},\cdots ,\vect{w}_{m}\right\} $
is said to be an
\index{orthonormal}\index{vectors!orthonormal}
 \textbf{orthonormal} set if 
\[
\vect{w}_i \dotprod \vect{w}_j = \delta _{ij} = \left\{ 
\begin{array}{c}
1\text{ if }i=j \\ 
0\text{ if }i\neq j
\end{array}
\right.
\]
\end{definition}

Note that all orthonormal sets are orthogonal, but the reverse is not
necessarily true since the vectors may not be normalized. In order to
normalize the vectors, we simply need divide each one by its length.

\begin{definition}{Normalizing an Orthogonal Set}{normalizing}
Normalizing an orthogonal set is the process of 
turning an orthogonal (but not orthonormal) set into
an orthonormal set.
If $\{ \vect{u}_1, \vect{u}_2, \ldots, \vect{u}_k\}$
is an orthogonal subset of $\mathbb{R}^n$, 
then 
\[ \left\{
\frac{1}{\vectlength\vect{u}_1\vectlength}\vect{u}_1,
\frac{1}{\vectlength\vect{u}_2\vectlength}\vect{u}_2, \ldots,
\frac{1}{\vectlength\vect{u}_k\vectlength}\vect{u}_k \right\}
\]
is an orthonormal set.
\end{definition}

We illustrate this concept in the following example. 

\begin{example}{Orthonormal Set}{orthonormalset}
Consider the set of vectors  given by 
\[
\left\{ \vect{u}_1, \vect{u}_2 \right\} = \left\{
\left[
\begin{array}{c}
1 \\
1 
\end{array}
\right], 
\left[
\begin{array}{r}
-1 \\
1
\end{array}
\right]
\right\}
\]
Show that it is an orthogonal set of vectors  but not an orthonormal one. Find the corresponding orthonormal set. 
\end{example}

\begin{solution}
One easily verifies that $\vect{u}_1 \dotprod \vect{u}_2 = 0$ and
$\left\{ \vect{u}_1, \vect{u}_2 \right\}$ is an orthogonal set of
vectors. On the other hand one can compute that $\vectlength \vect{u}_1 \vectlength= \vectlength \vect{u}_2 \vectlength =
\sqrt{2} \neq 1$ and thus it is not an orthonormal set.

Thus to find a corresponding orthonormal set, we simply need to
normalize each vector. We will write $\{ \vect{w}_1, \vect{w}_2 \}$
for the corresponding orthonormal set. Then,
\begin{eqnarray*}
\vect{w}_1 &=& \frac{1}{\vectlength \vect{u}_1 \vectlength} \vect{u}_1\\
&=& \frac{1}{\sqrt{2}} \left[
\begin{array}{c}
1 \\
1 
\end{array}
\right] \\
&=&
\leftB
\begin{array}{c}
\vspace{0.05in} \frac{1}{\sqrt{2}}\\
\vspace{0.05in}\frac{1}{\sqrt{2}} 
\end{array}
\rightB
\end{eqnarray*}

Similarly, 
\begin{eqnarray*}
\vect{w}_2 &=& \frac{1}{\vectlength \vect{u}_2 \vectlength} \vect{u}_2\\
&=& \frac{1}{\sqrt{2}} \left[
\begin{array}{r}
-1 \\
1 
\end{array}
\right] \\
&=&
\leftB
\begin{array}{r}
\vspace{0.05in} -\frac{1}{\sqrt{2}}\\
\vspace{0.05in}\frac{1}{\sqrt{2}} 
\end{array}
\rightB
\end{eqnarray*}

Therefore the corresponding orthonormal set is 
\[
\left\{ \vect{w}_1, \vect{w}_2 \right\} = 
\left\{
\leftB
\begin{array}{c}
\vspace{0.05in} \frac{1}{\sqrt{2}}\\
\vspace{0.05in}\frac{1}{\sqrt{2}} 
\end{array}
\rightB,
\leftB
\begin{array}{r}
\vspace{0.05in} -\frac{1}{\sqrt{2}}\\
\vspace{0.05in}\frac{1}{\sqrt{2}} 
\end{array}
\rightB
\right\} 
\]

You can verify that this set is orthogonal.
\end{solution}

Consider an orthogonal set of vectors in $\mathbb{R}^n$, written $\{
\vect{w}_1, \cdots, \vect{w}_k \}$ with $k \leq n$. The span of these
vectors is a subspace $W$ of $\mathbb{R}^n$. If we
could show that this orthogonal set is also linearly independent, we
would have a basis of $W$. We will show this in the next theorem.

\begin{theorem}{Orthogonal Basis of a Subspace}{orthbasis}
Let $ \{ \vect{w}_1, \vect{w}_2, \cdots, \vect{w}_k \}$ be an
orthonormal set of vectors in $\mathbb{R}^n$. Then this set is
linearly independent and forms a basis for the subspace $W =
\func{span} \{ \vect{w}_1, \vect{w}_2, \cdots, \vect{w}_k \}$.
\end{theorem}

\begin{proof}
To show it is a linearly independent set, suppose a linear combination
of these vectors equals $\vect{0}$, such as:
\[
a_1 \vect{w}_1 + a_2 \vect{w}_2 + \cdots + a_k \vect{w}_k = \vect{0}, a_i \in \mathbb{R} 
\]
We need to show that all $a_i = 0$. To do so, take the dot product of
each side of the above equation with the vector $\vect{w}_i$ and obtain the following. 

\begin{eqnarray*}
\vect{w}_i \dotprod (a_1 \vect{w}_1 + a_2 \vect{w}_2 + \cdots + a_k \vect{w}_k ) &=& \vect{w}_i \dotprod \vect{0}\\
a_1 (\vect{w}_i \dotprod \vect{w}_1) + a_2 (\vect{w}_i \dotprod \vect{w}_2) + \cdots + a_k (\vect{w}_i \dotprod \vect{w}_k)  &=& 0 
\end{eqnarray*}
Now since the set is orthogonal, $\vect{w}_i \dotprod \vect{w}_m = 0$ for all $m \neq i$, so we have:
\[
a_1 (0) + \cdots + a_i(\vect{w}_i \dotprod \vect{w}_i) + \cdots + a_k (0) = 0
\]
\[
a_i \vectlength \vect{w}_i \vectlength^2 = 0
\]

Since the set is orthogonal, we know that $\vectlength \vect{w}_i \vectlength^2  \neq 0$. It follows that $a_i =0$. Since the $a_i$ was chosen arbitrarily, the set $\{ \vect{w}_1, \vect{w}_2, \cdots, \vect{w}_k \}$ is linearly independent. 

Finally since $W = \mbox{span} \{ \vect{w}_1, \vect{w}_2, \cdots,
\vect{w}_k \}$, the set of vectors also spans $W$ and therefore forms a basis of $W$. 

\end{proof}

If an orthogonal set is a basis for a subspace, we call this an
orthogonal basis. Similarly, if an orthonormal set is a basis, we call
this an orthonormal basis.

We conclude this section with a discussion of Fourier expansions. Given any orthogonal basis $B$ of $\mathbb{R}^n$ and an arbitrary vector $\vect{x} \in \mathbb{R}^n$, how do we express $\vect{x}$ as a linear combination of vectors in $B$? The solution is Fourier expansion. 

\begin{theorem}{Fourier Expansion}{fourierexpansion}
Let $V$ be a subspace of $\mathbb{R}^n$ and suppose $\{ \vect{u}_1, \vect{u}_2, \ldots, \vect{u}_m \}$
is an orthogonal basis of $V$. 
Then for any $\vect{x}\in V$,
\[ \vect{x} = 
\left(\frac{\vect{x}\dotprod \vect{u}_1}{\vectlength \vect{u}_1 \vectlength^2}\right) \vect{u}_1 +
\left(\frac{\vect{x}\dotprod \vect{u}_2}{\vectlength \vect{u}_2 \vectlength^2}\right) \vect{u}_2 +
\cdots +
\left(\frac{\vect{x}\dotprod \vect{u}_m}{\vectlength \vect{u}_m \vectlength^2}\right) \vect{u}_m.
\]
This expression is called the Fourier expansion 
of $\vect{x}$, and 
\[ \frac{\vect{x}\dotprod \vect{u}_j}{\vectlength \vect{u}_j \vectlength^2},\]
$j=1,2,\ldots,m$
are the Fourier coefficients.
\end{theorem}

Consider the following example.

\begin{example}{Fourier Expansion}{fourier}
Let
$\vect{u}_1= \leftB\begin{array}{r} 1 \\ -1 \\ 2 \end{array}\rightB,
\vect{u}_2= \leftB\begin{array}{r} 0 \\ 2 \\ 1  \end{array}\rightB$,
and
$\vect{u}_3 =\leftB\begin{array}{r} 5 \\ 1 \\ -2 \end{array}\rightB$, 
and let 
$\vect{x} =\leftB\begin{array}{r} 1 \\ 1 \\ 1 \end{array}\rightB$.

Then $B=\{ \vect{u}_1, \vect{u}_2, \vect{u}_3\}$
is an orthogonal basis of $\mathbb{R}^3$. 

Compute the Fourier expansion of $\vect{x}$, thus writing $\vect{x}$ as  a linear combination of the vectors of $B$. 
\end{example}

\begin{solution}
Since $B$ is a basis (verify!) there is a unique way to express $\vect{x}$ as a
linear combination of the vectors of $B$. Moreover since $B$ is an
orthogonal basis (verify!), then this can be done by computing the
Fourier expansion of $\vect{x}$.

That is:

\[ 
\vect{x}   = 
\left(\frac{\vect{x}\dotprod \vect{u}_1}{\vectlength \vect{u}_1 \vectlength^2}\right) \vect{u}_1 +
\left(\frac{\vect{x}\dotprod \vect{u}_2}{\vectlength \vect{u}_2 \vectlength^2}\right) \vect{u}_2 +
\left(\frac{\vect{x}\dotprod \vect{u}_3}{\vectlength \vect{u}_3 \vectlength^2}\right) \vect{u}_3.
\]

We readily compute: 

\[
\frac{\vect{x}\dotprod\vect{u}_1}{\vectlength \vect{u}_1 \vectlength^2} = \frac{2}{6}, \; 
\frac{\vect{x}\dotprod\vect{u}_2}{\vectlength \vect{u}_2 \vectlength^2} = \frac{3}{5},
\mbox{ and }
\frac{\vect{x}\dotprod\vect{u}_3}{\vectlength \vect{u}_3 \vectlength^2} = \frac{4}{30}.\]

Therefore, 
\[ \leftB\begin{array}{r} 1 \\ 1 \\ 1 \end{array}\rightB
= \frac{1}{3}\leftB\begin{array}{r} 1 \\ -1 \\ 2 \end{array}\rightB
+\frac{3}{5}\leftB\begin{array}{r} 0 \\ 2 \\ 1  \end{array}\rightB
+\frac{2}{15}\leftB\begin{array}{r} 5 \\ 1 \\ -2 \end{array}\rightB.\]
\end{solution}