\todo[inline, color = yellow]{
\textbf{Major things to do:}
- Re-evaluate each section and timeline for completion.,Work on Section 3
- Write a sensitivity analysis section
- Create code and link it to examples
 - Build out the Excel and Python sections
- Decide what Linear Algebra Appendices to add
\textbf{Minor things to do:}
- Clean up preambles
- Standardize LP formatting
- Clean up metadata/references for all figures
- Fix Compile errors
- Add unnumbered "Resources and References" section to end of each chapter.
\textbf{Things completed: }
- Make slides template
}
\chapter*{Introduction}
\section*{Letter to instructors}
\todo[inline]{Fill in letter to professors and how to use the book}
This is an introductory book for students to learn optimization theory, tools, and applications.  The two main goals are (1) students are able to apply the of optimization in their future work or research (2) students understand the concepts underlying optimization solver and use this knowledge to use solvers effectively.

This book was based on a sequence of course at Virginia Tech in the Industrial and Systems Engineering Department.   The courses are \emph{Deterministic Operations Research I} and \emph{Deterministic Operations Reserach II}.   The first course focuses on linear programming, while the second course covers integer programming an nonlinear programming.  As such, the content in this book is meant to cover 2 or more full courses in optimization.

The book is designed to be read in a linear fashion. That said, many of the chapters are mostly independent and could be rearrranged, removed, or shortened as desited.  This is an open source textbook, so use it as you like.  You are encouraged to share adaptations and improvements so that this work can evolve over time.




\section*{Letter to students}



This book is designed to be a resource for students interested in learning what optimizaiton is, how it works, and how you can apply it in your future career.  The main focus is being able to apply the techniques of optimization to problems using computer technology while understanding (at least at a high level) what the computer is doing and what you can claim about the output from the computer.

Quite importantly, keep in mind that when someone claims to have \emph{optimized} a problem, we want to know what kind of gaurantees they have about how good their solution is.  Far too often, the solution that is provided is suboptimal by 10\%, 20\%, or even more.  This can mean spending excess amounts of money, time, or energy that could have been saved. And when problems are at a large scale, this can easily result in millions of dollars in savings.

For this reason, we will learn the perspective of \emph{mathematical programming} (a.k.a. mathematical optimization).  The key to this study is that we provide gaurantees on how good a solution is to a given problem.  We will also study how difficult a problem is to solve.  This will help us know (a) how long it might take to solve it and (b) how good a of a solution we might expect to be able to find in reasonable amount of time. We will later study \emph{heuristic} methods - these methods typically do not come with gaurantees, but tend to help find quality solutions.

Note: Although there is some computer programming required in this work, this is not a course on programming.   Thanks to the fantastic modelling packages available these days, we are able to solve complicated problems with little programming effort.   The key skill we will need to learn is \emph{mathematical modeling}: converting words and ideas into numbers and variables in order to communicate problems to a compter so that it can solve a problem for you.  

As a main element of this book, we would like to make the process of using code and software as easy as possible.  Attached to most examples in the book, there will be links to code that implements and solves the problem using several different tools from Excel and Python.  These examples can should make it easy to solve a similar problem with different data, or more generally, can serve as a basis for solving related problems with similar structure.

\subsubsection*{How to use this book}

\emph{Skim ahead.} We recommend that before you come across a topic in lecture, that you skim the relevant sections ahead of time to get a broad overview of what is to come.   This may take only a fraction of the time that it may take for you to read it.

\emph{Read the expected outcomes.} At the beginning of each section, there will be a list of expected outcomes from that section.  Review these outcomes before reading the section to help guide you through what is most relevant for you to take away from the material.  This will also provide a brief look into what is to follow in that section.

\emph{Read the text.} Read carefully the text to understand the problems and techniques.  We will try to provide a plethora of examples, therefore, depending on your understanding of a topic, you many need to go carefully over all of the examples.

\emph{Explore the resources.} Lastly, we recognize that there are many alternative methods of learning given the massive amounts of information and resources online.   Thus, at the end of each section, there will be a number of superb resources that available on the internet in different formats.   There are other free textbooks, informational websites, and also number of fantastic videos posted to youtube.  We encourage to explore the resources to get another perspective on the material or to hear/read it taught from a differnt point of view or in presentation style.  

\todo[inline]{Add how to use example boxes and complexity boxes}

\subsubsection*{Outline of this book}
This book is divided in to 4 Parts: 
\todo[inline]{Parts are not left aligned????}
\begin{enumerate}
\item[Part I] Linear Programming,
\item[Part II] Integer Programming,
\item[Part III] Discrete Algorithms,
\item[Part IV]  Nonlinear Programming.
\end{enumerate}
There are also a number of chapters of background material in the Appendix.

The content of this book is designed to encompass 2-3 full semester courses in an industrial engineering department. 

\subsubsection*{Work in progress}
This book is still a work in progress, so please feel free to send feedback, questions, comments, and edits to Robert Hildebrand at \url{open.optimization@gmail.com}.

\todo[inline]{Change hyperlink for website.    make them just blue - brighter than headings.}