\chapter{Software - Python}
\todoChapter{ {\color{gray}10\% complete. Goal 80\% completion date: August 20}\\
Notes: }
\begin{outcome}
\begin{itemize}
\item Install and get python up and running in some form
\item Introduce basic python skills that will be helpful
\end{itemize}
\end{outcome}

\begin{resource}
\begin{itemize}
\item\href{https://open.umn.edu/opentextbooks/textbooks/a-byte-of-python}{A Byte of Python}
\item\href{https://github.com/swaroopch/byte-of-python}{Github - Byte of Python} (CC-BY-SA)
\end{itemize}
\end{resource}

\subimport{./foundationsAppliedMathematicsLabs/Appendices/Installation/}{Installation} %Chapter
\subimport{./foundationsAppliedMathematicsLabs/Appendices/NumpyVisualGuide/}{NumpyVisualGuide} %Chapter
\subimport{./foundationsAppliedMathematicsLabs/Appendices/MatplotlibCustomization/}{MatplotlibCustomization} %Chapter

\section{Networkx - A Python Graph Algorithms Package}


\section{PuLP - An Optimization Modeling Tool for Python}
\begin{outcome}
\begin{itemize}
\item Install and import PuLP
\item Run basic first PuLP model
\item Run "advanced" PuLP model using the algebraic modeling approach and importing data.
\item Explore PuLP objects and possibilities
\item Solve a Multi-Objective problem
\end{itemize}
\end{outcome}


\begin{resource}
\begin{itemize}
\item \href{https://coin-or.github.io/pulp/}{Documentation}
\item \href{https://pypi.org/project/PuLP/}{PyPi installation}
\item \href{https://github.com/coin-or/pulp/tree/master/examples}{Examples}
\item \href{https://benalexkeen.com/linear-programming-with-python-and-pulp-part-1/}{Blog with tutorial}
\end{itemize}
\end{resource}

PuLP is an optimizaiton modeling language that is written for Python.  It is free and open source.  Yay!   See Section ?? for a discussion of other options for implementing your optimization problem.
PuLP is convenient for it's simple syntax and easy installation.   
 
Key benefits of using an algebraic modeling language like PuLP over Excel
\begin{itemize}
\item Easily readable models
\item Precompute parameters within Python
\item Reuse of common optimization models without recreating the equations
\end{itemize}

We will follow the introduction to pulp Jupyter Notebook Tutorial and the following application with a cleaner implementation.

\input{pulp_import.tex}
\input{pulp_transportation.tex}

\input{transportation-multi-objective/transportation-multi-objective-jupyter.tex}
\section{Jupyter Notebooks}
\begin{resource}
\begin{itemize}
\item \url{https://github.com/mathinmse/mathinmse.github.io/blob/master/Lecture-00B-Notebook-Basics.ipynb}
\item \url{https://github.com/mathinmse/mathinmse.github.io/blob/master/Lecture-00C-Writing-In-Jupyter.ipynb}
\end{itemize}
\end{resource}

\section{Reading and Writing}
\url{https://github.com/mathinmse/mathinmse.github.io/blob/master/Lecture-10B-Reading-and-Writing-Data.ipynb}

\section{Python Crash Course}
\url{https://github.com/rpmuller/PythonCrashCourse}


\section{Gurobi}
You can find lots great tutorial vidoes through Gurobi's youtube chanel.  

\todoSection{Link to introductory Gurobi files and installation instructions.
Show how grblogtools can show progress of integer programs and compare models across a set of instances.}

\subsection{Introductory Gurobi Examples}
\begin{enumerate}
\item \href{https://github.com/open-optimization/open-optimization-or-examples/blob/master/gurobi_helper_files/Gurobi-first-model.ipynb}{Gurobi-first-model.ipynb}
\item \href{https://github.com/open-optimization/open-optimization-or-examples/blob/master/gurobi_helper_files/Gurobi-knapsack1.ipynb}{Gurobi-knapsack1.ipynb}
\item \href{https://github.com/open-optimization/open-optimization-or-examples/blob/master/gurobi_helper_files/Gurobi-knapsack2.ipynb}{Gurobi-knapsack2.ipynb}
\item \href{https://github.com/open-optimization/open-optimization-or-examples/blob/master/gurobi_helper_files/Gurobi-knapsack3-Copy1.ipynb}{Gurobi-knapsack3.ipynb}
\item \href{https://github.com/open-optimization/open-optimization-or-examples/blob/master/gurobi_helper_files/Gurobi-knapsack-SOS1.ipynb}{Gurobi-knapsack-SOS1.ipynb}
\item \href{https://github.com/open-optimization/open-optimization-or-examples/blob/master/gurobi_helper_files/Gurobi-SOS2-piecewise-linear.ipynb}{Gurobi-SOS2-piecewise-linear.ipynb}
\end{enumerate}

\href{https://github.com/Gurobi}{Gurobi on GitHub}


\href{https://github.com/Gurobi/modeling-examples}{Gurobi Modeling Examples}

\href{https://github.com/Gurobi/grblogtools}{Gurobi Log Tools}


\section{Plots, Pandas, and Geopandas}
\todoSection{}
\subsection{Matplotlib.pyplot}
This is perhaps the most commonly used package for plotting.   The functionality is a lot like MatLab's plotting functionality.

\todo[inline]{Add Matplotlib tutorial reference.}


\subsection{Pandas}
Pandas is the most used package for manupulating data in python.   You can load data from excel, csv, json, or other types of files, and manipulate data in a variety of ways.
\todo[inline]{Add Panadas tutorial.}

\subsection{Geopandas}
Geopandas is a fantastic python package that allows you to interact with geospatial data, make plots, and analyze data.   This typically involves loading shapefiles that describe a region (or many regions) into a geopandas dataframe.  These function much like a regular pandas dataframe, but has a extra column for the geometry of the data and has other functionality for these types of dataframes.  

Here are some great tutorials.
\begin{itemize}
\item \href{https://jcutrer.com/python/learn-geopandas-plotting-usmaps}{tutorial}
\item \href{https://github.com/joncutrer/geopandas-tutorial}{tutorial github}
\end{itemize}
\section{Google OR Tools}
\todoSection{Give introduction to Google OR Tools and what problems it can solve.}
Google has been building their own set of optimization tools and marketing them to solve operations research problems.   They have a variety of type of solvers and have a number of example problems where they use the appropriate solver to solve a problem.   This is a constantly evolving set of tool. See the link below examples on vehicle routing, scheduling,  and more.  


\begin{center}
\fbox{
\href{https://developers.google.com/optimization}{Google OR Tools}}
\end{center}

\href{https://www.youtube.com/watch?v=AJ6LeiMe_PQ&t=757s&ab_channel=MixedIntegerProgramming}{Youtube!}
\href{https://www.youtube.com/watch?v=iF2rHY318iU&ab_channel=MixedIntegerProgramming}{Video!}

