\subsubsection*{How to use this book}

\emph{Skim ahead.} We recommend that before you come across a topic in lecture, that you skim the relevant sections ahead of time to get a broad overview of what is to come.   This may take only a fraction of the time that it may take for you to read it.

\emph{Read the expected outcomes.} At the beginning of each section, there will be a list of expected outcomes from that section.  Review these outcomes before reading the section to help guide you through what is most relevant for you to take away from the material.  This will also provide a brief look into what is to follow in that section.

\emph{Read the text.} Read carefully the text to understand the problems and techniques.  We will try to provide a plethora of examples, therefore, depending on your understanding of a topic, you many need to go carefully over all of the examples.

\emph{Explore the resources.} Lastly, we recognize that there are many alternative methods of learning given the massive amounts of information and resources online.   Thus, at the end of each section, there will be a number of superb resources that available on the internet in different formats.   There are other free textbooks, informational websites, and also number of fantastic videos posted to youtube.  We encourage to explore the resources to get another perspective on the material or to hear/read it taught from a differnt point of view or in presentation style.  

\todo[inline]{Add how to use example boxes and complexity boxes}