\chapter*{Preface}

This book is a living document.  Feedback, questions, comments, and edits are welcome at \url{open.optimization@gmail.com}.

This book is being written and compiled using a number of open source materials.   We will  strive to properly cite all resources used and give references on where to find these resources.  Although the material used in this book comes from a variety of licences, everything used here will be CC-BY-SA 4.0 compatible, and hence, the entire book will fall under a CC-BY-SA 4.0 license.

\begin{center}
\fontsize{14pt}{16pt}\selectfont\textcolor{titletextcolour}{\textbf{MAJOR ACKNOWLEDGEMENTS}}
\end{center}

I would like to acknowledge that substantial parts of this book were adapted from the following open-source materials.
All content used is compatible with the CC~BY-SA~4.0 license of this book.

\begin{itemize}
\item \textbf{``A First Course in Linear Algebra''} by Lyryx Learning (based on original text by Ken Kuttler), licensed under \textbf{CC~BY~4.0}.   A majority of their formatting was used along with selected sections that make up the appendix sections on linear algebra. We are extremely grateful to Lyryx for sharing their files with us.  They do an amazing job compiling their books, and the templates and formatting that we have borrowed here clearly took a lot of work to set up.  Thank you for sharing all of this material to make structuring and formatting this book much easier! See subsequent page for list of contributors.

\item \textbf{``Foundations of Applied Mathematics''} with many contributors, licensed under \textbf{CC~BY~3.0~US}.  See \url{https://github.com/Foundations-of-Applied-Mathematics}.  Several sections from these notes were used along with some formatting.  Some of this content has been edited or rearranged to suit the needs of this book.  This content comes with some great references to code and nice formatting to present code within the book.   See subsequent page with list of contributors.

\item \textbf{``Linear Inequalities and Linear Programming''} by
Kevin Cheung, licensed under \textbf{CC~BY-SA~4.0}.  See \url{https://github.com/dataopt/lineqlpbook}.  These notes are posted on GitHub in a ``.Rmd'' format for nice reading online.  This content was converted to \LaTeX\ using Pandoc.    These notes make up a substantial section of the Linear Programming part of this book.

\item \textbf{Linear Programming notes} by Douglas Bish.  These notes also make up a substantial section of the Linear Programming part of this book.

\item \textbf{``Applied Finite Mathematics''} by Rupinder Sekhon and Roberta Bloom, licensed under \textbf{CC~BY~4.0}.  See \url{https://math.libretexts.org/Bookshelves/Applied_Mathematics/Applied_Finite_Mathematics_(Sekhon_and_Bloom)}.  Content from Chapter~1 (Linear Equations) was adapted for the appendix on linear equations.
\end{itemize}

I would also like to acknowledge Laurent Porrier and Diego Moran for contributing various notes on linear and integer programming.

I would also like to thank Jamie Fravel for helping to edit this book and for contributing chapters, examples, and code.

\newpage

