\chapter*{Sources and Attribution}
\addcontentsline{toc}{chapter}{Sources and Attribution}
\markboth{Sources and Attribution}{Sources and Attribution}

This book adapts material from several open-source textbooks and lecture
notes. Every source listed here uses a license that is compatible with
the Creative Commons Attribution-ShareAlike~4.0 International
(CC~BY-SA~4.0) license under which this book is distributed. Each entry
records the original author, source URL, original license, and how the
material was used. Modifications to all adapted material were made by
Robert Hildebrand and other contributors to this book.

%-------------------------------------------------------------------
\section*{Adapted Open-Source Textbooks and Notes}
%-------------------------------------------------------------------

\paragraph{A First Course in Linear Algebra}\hspace{0pt}\\[-0.6em]
\begin{description}
  \item[Author:] Ken Kuttler, with adaptations by Lyryx Learning
  \item[Source:] \url{https://lyryx.com/first-course-linear-algebra/}
  \item[License:] Creative Commons Attribution 4.0 International (CC~BY~4.0)
  \item[Local copy:] \texttt{external-sources/aFirstCourseLinearAlgebra/}
  \item[Used in:] Appendix on Linear Algebra (vectors, dot product,
    cross product, projections, Gram--Schmidt, planes, parametric
    lines). The book also reuses Lyryx's chapter and theorem
    formatting templates.
  \item[Compatibility:] CC~BY~4.0 material may be incorporated into a
    CC~BY-SA~4.0 work with attribution. The adapted material here is
    relicensed under CC~BY-SA~4.0 as part of this book.
\end{description}

\paragraph{Foundations of Applied Mathematics: Volume 2 -- Algorithms,
  Approximation, and Optimization}\hspace{0pt}\\[-0.6em]
\begin{description}
  \item[Authors:] Jeffrey Humpherys, Tyler Jarvis, and contributors at
    Brigham Young University (full contributor list in
    \texttt{frontmatter/contributors.tex})
  \item[Source:] \url{https://github.com/Foundations-of-Applied-Mathematics}
  \item[License:] Creative Commons Attribution 3.0 United States License
    (CC~BY~3.0~US)
  \item[Local copy:] \texttt{external-sources/foundationsAppliedMathematicsLabs/}
  \item[Used in:] Selected sections, code formatting templates, and
    references to associated lab code throughout the book.
  \item[Compatibility:] CC~BY~3.0 permits adaptation under any
    compatible license; the adapter's license here is CC~BY-SA~4.0.
    Attribution is retained for all adapted material.
\end{description}

\paragraph{Linear Inequalities and Linear Programming}\hspace{0pt}\\[-0.6em]
\begin{description}
  \item[Author:] Kevin Cheung, Carleton University
  \item[Source:] \url{https://github.com/dataopt/lineqlpbook}
  \item[License:] Creative Commons Attribution-ShareAlike 4.0
    International (CC~BY-SA~4.0)
  \item[Local copy:] \texttt{external-sources/lineqlpbook/}
  \item[Used in:] Substantial portions of Part~I (Linear Programming):
    notation, basic feasible solutions, the Fundamental Theorem of LP,
    Farkas' lemma, complementary slackness, duality, the Fourier--Motzkin
    elimination method, and the graphical example. License is documented
    via headers in every \texttt{.tex} source file.
  \item[Compatibility:] CC~BY-SA~4.0 source content is directly
    compatible with CC~BY-SA~4.0 adaptation; no version transition.
\end{description}

\paragraph{Optimization Notes (Linear Programming Notes)}\hspace{0pt}\\[-0.6em]
\begin{description}
  \item[Author:] Douglas Bish, Virginia Tech
  \item[License:] Creative Commons Attribution-ShareAlike 4.0
    International (CC~BY-SA~4.0)
  \item[Local copy:] \texttt{external-sources/Opt\_Notes/}
  \item[Used in:] Substantial portions of Part~I (Linear Programming),
    including modeling, the simplex method exposition, and sensitivity
    analysis. License is documented in
    \texttt{external-sources/Opt\_Notes/README.md}.
  \item[Compatibility:] CC~BY-SA~4.0 source content is directly
    compatible with CC~BY-SA~4.0 adaptation.
\end{description}

\paragraph{Applied Finite Mathematics, Chapter 1: Linear Equations}\hspace{0pt}\\[-0.6em]
\begin{description}
  \item[Authors:] Rupinder Sekhon and Roberta Bloom
  \item[Source:] \url{https://math.libretexts.org/Bookshelves/Applied_Mathematics/Applied_Finite_Mathematics_(Sekhon_and_Bloom)}
  \item[License:] Creative Commons Attribution 4.0 International
    (CC~BY~4.0), via LibreTexts
  \item[Local copy:] \texttt{external-sources/applied-finite-mathematics/}
    (with LICENSE file)
  \item[Used in:] Appendix A (Linear Equations), including text and
    figures adapted from LibreTexts Chapter~1.
  \item[Compatibility:] CC~BY~4.0 material may be incorporated into a
    CC~BY-SA~4.0 work with attribution. A visible attribution paragraph
    appears at the start of the appendix chapter.
\end{description}

%-------------------------------------------------------------------
\section*{Image Attributions}
%-------------------------------------------------------------------

\paragraph{Bridges of K\"onigsberg.}
The historical engraving of the seven bridges (used in the Graph
Theory chapter) is in the public domain. The line-drawing variant
included here is by Bogdan Giusca (Wikipedia user Bogdang) and is
licensed under Creative Commons Attribution-ShareAlike~3.0 (CC~BY-SA~3.0).
Source: \url{https://commons.wikimedia.org/wiki/File:Konigsberg_bridges.png}.

\paragraph{All other figures.}
All TikZ figures, plots, and diagrams in this book are original works
by Robert Hildebrand and contributors, created specifically for this
text. Earlier versions of some Wikipedia-derived figures
(e.g., generic linear-programming and integer-programming illustrations)
have been replaced with original TikZ versions. The four
multi-objective optimization figures in Chapter~13 are original
works by Robert Hildebrand.

%-------------------------------------------------------------------
\section*{Other Acknowledgements}
%-------------------------------------------------------------------

We thank Laurent Poirrier and Diego Moran for contributing notes on
linear and integer programming, and Jamie Fravel for editing,
contributing chapters, and contributing example code. We also thank
Irma Adams for researching and writing the ``pioneer spotlight''
biographical sidebars that appear throughout the book (including the
profiles of George Dantzig and other founders of the field). Joshua
Emmanuel provided early guidance on graphical-method exposition during
course development; the current text on the graphical method is
original to this book.

%-------------------------------------------------------------------
\section*{License Compatibility Summary}
%-------------------------------------------------------------------

\begin{center}
\renewcommand{\arraystretch}{1.25}
\begin{tabular}{l l l}
\toprule
\textbf{Source license} & \textbf{Compatibility} & \textbf{Mechanism} \\
\midrule
CC~BY-SA~4.0 & Direct & Same license; attribution retained \\
CC~BY~4.0    & Direct & Adapted into CC~BY-SA~4.0 with attribution \\
CC~BY~3.0~US & Indirect & Adapter's-license provision permits CC~BY-SA~4.0 \\
Public Domain & Direct & No restriction; attribution provided \\
\bottomrule
\end{tabular}
\end{center}

\noindent The full text of the CC~BY-SA~4.0 license under which this
book is distributed is available at \url{https://creativecommons.org/licenses/by-sa/4.0/}
and in the file \texttt{LICENSE-Content} in the book's source repository.
Code released with this book is licensed separately under the
\textbf{MIT License}, with text in \texttt{LICENSE-Code}.

\cleardoublepage
